---
title: "Data Curation and Correction"
author: "Marco Fioriani, Sara Leka, Carlo Rossetto"
date: "2025-11-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(dplyr)
  library(biomaRt)
  library(sva)
  library(pheatmap) 
  library(ggplot2)
})
```

##Introduction

This document covers Part 1 of our analysis: Data Curation, Quality Control, and Correction.

###Goals:
- Load and format our aggregated count and TPM matrices.
- Create and load our metadata file.
- Filter out non-expressed genes (as per Tutorial.pdf, Sec 3-3).
- Check for hematopoietic (immune cell) contamination, which is a known risk of the FACS-sorting sample prep (as per Krausgruber et al. 2020).
- Use ComBat_seq to correct for batch/individual effects, which also helps mitigate the contamination (as per Tutorial.pdf, Sec 3-3).
- Save the final, clean matrices for the next parts of our analysis.

##1. Loading and preparing data
```{r}
counts_data = read.csv("final_counts_matrix.csv")
tpm_data = read.csv("final_tpm_matrix.csv")
meta_data = read.csv("SraRunTable.csv")

prep_matrix = function(df) {
  df_prepped = df %>%
  column_to_rownames(var = "gene_id")
  return(df_prepped)
}

counts_matrix = prep_matrix(counts_data)
tpm_matrix = prep_matrix(tpm_data)
```

##2. Filter lowly-expressed genes
```{r}
avg_expr = rowMeans(tpm_matrix)
layout(matrix(1:2, nrow=1))
hist(avg_expr, main = "Histogram of Average TPM", xlab = "Average TPM")
hist(log10(avg_expr+ 1), main = "Histogram of Log10(TPM+1)", xlab = "Log10(TPM+1)")
```

```{r}
ggplot(data.frame(avg_expr), aes(x=avg_expr)) +
  geom_histogram(bins = 50) +
  scale_x_continuous(breaks = c(0,1,10,100,1000,10000,20000), trans="log1p",
expand=c(0,0)) +
  scale_y_continuous(breaks = c(0, 10, 100, 1000, 10000), trans="log1p") +
  theme_minimal() +
  ggtitle("Distribution of Average Gene Expression (Log scale)")
```

```{r}
num_det <- rowSums(tpm_matrix > 0)
hist(num_det, main = "Histogram of Gene Detections", xlab = "Number of Samples Gene is Detected In")
```



```{r}
expressed = rowMeans(tpm_matrix > 0) >= 0.5 | rowMeans(tpm_matrix) >= 1
print(paste("Original gene count:", nrow(tpm_matrix)))

tpm_matrix_filtered = tpm_matrix[expressed, ]
counts_matrix_filtered = counts_matrix[expressed, ]

print(paste("New filtered gene count:", nrow(tpm_matrix_filtered)))
```

##3. FACS contamination check (by detectionb of the pan-immune cell marker Ptprc)

3a. Using biomaRt to switch from MGI gene_id symbols to common gene symbols
```{r}
# connecting to gene mouse database
mgi_gene_ids = rownames(tpm_matrix_filtered)
ensembl <- useMart("ensembl", dataset = "mmusculus_gene_ensembl")

#creating a translation map 
gene_map <- getBM(
  attributes = c("mgi_symbol", "external_gene_name"), 
  filters = "mgi_symbol",
  values = mgi_gene_ids,         
  mart = ensembl
)

#finding the ptprc gene in our mgi gene_id
ptprc = gene_map$mgi_symbol[gene_map$external_gene_name == "Ptprc"]




# Check 1: Did we find the Ptprc gene in the biomaRt map?
if (length(ptprc) > 0 && !is.na(ptprc)) {
  
  # Check 2: Does that gene ID *also* exist in our *filtered* matrix?
  # (It could have been filtered out for low expression)
  if (ptprc %in% rownames(tpm_matrix_filtered)) {
    
    # 5. Extract Ptprc expression and join with metadata
    ptprc_expression <- tpm_matrix_filtered[ptprc, ] %>%
      t() %>%
      as.data.frame() %>%
      rownames_to_column(var = "Run") %>%
      left_join(meta_data, by = "Run")
    
    colnames(ptprc_expression)[2] <- "Ptprc_TPM"

    # 6. Plot the contamination
    print("Plotting expression of Ptprc (CD45) as a contamination check:")
    
    print( # We explicitly print the ggplot object inside the 'if' block
      ggplot(ptprc_expression, aes(x = Organ, y = Ptprc_TPM, fill = Organ)) +
        geom_boxplot() +
        geom_jitter(width = 0.1) +
        labs(
          title = "FACS Contamination Check: Ptprc (CD45) Expression",
          subtitle = "Ptprc is an immune cell marker and should be low in pure fibroblast samples.",
          y = "Ptprc Expression (TPM)",
          x = "Organ"
        ) +
        theme_bw() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
    )
    
  } else {
    print(paste("NOTE: 'Ptprc' (ID:", ptprc, ") was found in biomaRt but was FILTERED OUT for low expression."))
  }
  
} else {
  print("NOTE: 'Ptprc' gene not found after mapping with biomaRt.")
  print("This is OK, but it means we can't use it as our only QC check.")
  print("We will proceed with ComBat_seq correction as planned.")
}
```
```{r}
# --- ADD THIS CODE to the end of your "check-contamination" chunk ---

contamination_table <- ptprc_expression %>%
  filter(Organ %in% c("Kidney", "Spleen")) %>%
  select(Run, Organ, BioSample, Ptprc_TPM) %>%
  arrange(desc(Ptprc_TPM)) # Sort by highest contamination

knitr::kable(contamination_table, caption = "Ptprc (CD45) Contamination per Sample")
```
##4. Handling FACS contamination
###...

##5. PCA
```{r}
pca = prcomp(log1p(t(tpm_matrix_filtered)), center = TRUE, scale. = TRUE)
eigs = pca$sdev^2
plot(1:length(eigs), eigs)
```
```{r}
ggplot(data.frame(pca$x, meta_data)) +
  geom_point(aes(x = PC1, y = PC2, color = Organ), size = 5)
```
```{r}
write.csv(counts_matrix_filtered, "final_counts_filtered.csv", row.names = TRUE)

write.csv(tpm_matrix_filtered, "final_tpm_filtered.csv", row.names = TRUE)

write.csv(gene_map, "mgi_to_genesymbol_map.csv", row.names = FALSE)


print("Phase 1 complete. Files saved:")
print("- final_counts_filtered.csv (Use this for DESeq2)")
print("- final_tpm_filtered.csv (For visualizations)")
print("- mgi_to_genesymbol_map.csv (The gene ID map)")
```


