---
title: "Data Curation and Correction"
author: "Marco Fioriani, Sara Leka, Carlo Rossetto"
date: "2025-11-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(dplyr)
  library(biomaRt)
  library(sva)
  library(pheatmap) 
  library(ggplot2)
})
```

##Introduction

This document covers Part 1 of our analysis: Data Curation, Quality Control, and Correction.

###Goals:
- Load and format our aggregated count and TPM matrices.
- Create and load our metadata file.
- Filter out non-expressed genes (as per Tutorial.pdf, Sec 3-3).
- Check for hematopoietic (immune cell) contamination, which is a known risk of the FACS-sorting sample prep (as per Krausgruber et al. 2020).
- Use ComBat_seq to correct for batch/individual effects, which also helps mitigate the contamination (as per Tutorial.pdf, Sec 3-3).
- Save the final, clean matrices for the next parts of our analysis.

##1. Loading and preparing data
```{r}
counts_data = read.csv("final_counts_matrix.csv")
tpm_data = read.csv("final_tpm_matrix.csv")
meta_data = read.csv("SraRunTable.csv")

prep_matrix = function(df) {
  df_prepped = df %>%
  column_to_rownames(var = "gene_id")
  return(df_prepped)
}

counts_matrix = prep_matrix(counts_data)
tpm_matrix = prep_matrix(tpm_data)
```

##2. Filter lowly-expressed genes
```{r}
avg_expr = rowMeans(tpm_matrix)
layout(matrix(1:2, nrow=1))
hist(avg_expr)
hist(log10(avg_expr+ 1))
```

```{r}
ggplot(data.frame(avg_expr), aes(x=avg_expr)) +
  geom_histogram(bins = 50) +
  scale_x_continuous(breaks = c(0,1,10,100,1000,10000,20000), trans="log1p",
expand=c(0,0)) +
  scale_y_continuous(breaks = c(0,1), expand=c(0,0), trans="log1p") +
  theme_minimal()
```

```{r}
num_det <- rowSums(expr > 0)
hist(num_det)
```



```{r}
expressed = rowMeans(expr > 0) >= 0.5 | rowMeans(expr) >= 1
print(paste("Old gene count:", nrow(tpm_data)))

tpm_data = tpm_data[which(expressed),]
counts_data = counts_data[which(expressed),]
meta_data = meta_data[which(expressed),]

tpm_threshold <- 1

print(paste("New gene count:", nrow(tpm_data)))

```

