---
title: "Data Curation and Correction"
author: "Marco Fioriani, Sara Leka, Carlo Rossetto"
date: "2025-11-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(dplyr)
  library(biomaRt)
  library(sva)
  library(pheatmap) 
  library(ggplot2)
  library(gridExtra)
})
```

## Introduction

This document covers Part 1 of our analysis: Data Curation, Quality
Control, and Correction.

### Goals: - Load and format our aggregated count and TPM matrices. 
- Create and load our metadata file. 
- Filter out non-expressed genes (as per Tutorial.pdf, Sec 3-3). 
- Check for hematopoietic (immune cell) contamination, which is a known risk of the FACS-sorting sample prep (as per Krausgruber et al. 2020). - Use ComBat_seq to correct for batch/individual effects, which also helps mitigate the contamination (as per Tutorial.pdf, Sec 3-3). 
- Save the final, clean matrices for the next parts of our analysis.

## 1. Loading and preparing data
```{r}
counts_data = read.csv("final_counts_matrix.csv")
tpm_data = read.csv("final_tpm_matrix.csv")
meta_data = read.csv("SraRunTable.csv")

prep_matrix = function(df) {
  df_prepped = df %>%
  column_to_rownames(var = "gene_id")
  return(df_prepped)
}

counts_matrix = prep_matrix(counts_data)
tpm_matrix = prep_matrix(tpm_data)
```

## 2. Filtering out lowly-expressed genes
```{r}
avg_expr = rowMeans(tpm_matrix)
layout(matrix(1:2, nrow=1))
hist(avg_expr, main = "Histogram of Average TPM", xlab = "Average TPM")
hist(log10(avg_expr+ 1), main = "Histogram of Log10(TPM+1)", xlab = "Log10(TPM+1)")
```
```{r}
ggplot(data.frame(avg_expr), aes(x=avg_expr)) +
  geom_histogram(bins = 50) +
  scale_x_continuous(breaks = c(0,1,10,100,1000,10000,20000), trans="log1p",
expand=c(0,0)) +
  scale_y_continuous(breaks = c(0, 10, 100, 1000, 10000), trans="log1p") +
  theme_minimal() +
  ggtitle("Distribution of Average Gene Expression (Log scale)")
```

```{r}
num_det <- rowSums(tpm_matrix > 0)
hist(num_det, main = "Histogram of Gene Detections", xlab = "Number of Samples Gene is Detected In")
```
Filters for genes expressed in at least 50% of the samples or with an average expression of at least 1 TPM (saves genes that might be rare so detected in < 50% of samples but are highly expressed when they are present, e.g. a marker gene specific to just one organ)
```{r}
expressed = rowMeans(tpm_matrix > 0) >= 0.5 | rowMeans(tpm_matrix) >= 1
print(paste("Original gene count:", nrow(tpm_matrix)))
tpm_matrix_filtered = tpm_matrix[expressed, ]
counts_matrix_filtered = counts_matrix[expressed, ]

print(paste("New filtered gene count:", nrow(tpm_matrix_filtered)))
```

## 3. FACS contamination check (by detection of 5 immune cellmarkers Ptprc, Cd3e, Cd19, Adgre1, Itgam)

3a. Using biomaRt to switch from MGI gene_id symbols to common gene symbols
```{r}
# connecting to gene mouse database
immune_markers = c("Ptprc", "Cd3e", "Cd19", "Adgre1", "Itgam")
mgi_gene_ids = rownames(tpm_matrix_filtered)
ensembl = useMart("ensembl", dataset = "mmusculus_gene_ensembl")

#creating a translation map 
gene_map <- getBM(
  attributes = c("mgi_symbol", "external_gene_name"), 
  filters = "mgi_symbol",
  values = mgi_gene_ids,         
  mart = ensembl
)
```

3b. Bar plotting the contamination per organ per immune gene marker (Cd3e was not present in any sample so only 4 immune markers are displayed)
```{r}
plot_marker = function(gene_symbol, map, data, meta) {
  gene_id = map$mgi_symbol[map$external_gene_name == gene_symbol]
  gene_id = unique(gene_id)
  if (length(gene_id) > 0 && gene_id %in% rownames(data)) {
    expr_df = data[gene_id, ] %>%
      t() %>%
      as.data.frame() %>%
      rownames_to_column(var = "Run") %>%
      left_join(meta, by = "Run")
    colnames(expr_df)[2] = "TPM"
    
    p = ggplot(expr_df, aes(x = Organ, y = TPM, fill = Organ)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(width = 0.1, size = 1) +
      labs(title = gene_symbol, x = NULL, y = "TPM") +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
    return(p)
  } else {
    return(NULL)
  }
}

plot_list = list()
for (marker in immune_markers) {
  p = plot_marker(marker, gene_map, tpm_matrix_filtered, meta_data)
  if (!is.null(p)) {
    plot_list[[marker]] = p
  }
}

print("Plotting immune marker panel:")
grid.arrange(grobs = plot_list, ncol = 2)
```

##4. Handling FACS contamination (it will be used as a covariate in the DE design)

4a.Adding the averaged contamination score per sample to the metadata
```{r}
immune_ids = gene_map$mgi_symbol[gene_map$external_gene_name %in% immune_markers]
immune_expr = tpm_matrix_filtered[immune_ids, , drop = FALSE]

contamination_scores = colMeans(log1p(immune_expr))

score_df = data.frame(Run = names(contamination_scores), ContaminationScore = contamination_scores)
meta_data = meta_data %>% left_join(score_df, by = "Run")
print("Contamination scores calculated and added to metadata.")
print(head(meta_data[, c("Run", "Organ", "ContaminationScore")]))
```
4b. Visualization of the contamination
```{r}
immune_table = as.data.frame(t(immune_expr)) %>% rownames_to_column(var = "Run")
name_swap = setNames(gene_map$external_gene_name, gene_map$mgi_symbol)
colnames(immune_table)[-1] = name_swap[colnames(immune_table)[-1]]
immune_table_final = meta_data %>% 
  select(Run, Organ) %>% 
  right_join(immune_table, by = "Run") %>% 
  arrange(Organ)  
print(head(immune_table_final, 10))
  
write.csv(immune_table_final, "immune_marker_expression_per_sample.csv", row.names = FALSE)
```
```{r}
immune_expr_log = log1p(immune_expr)

annotation_col = meta_data %>% select(Run, Organ) %>% column_to_rownames("Run")

pheatmap(
  immune_expr_log,
  main = "Immune Marker Expression (Log1p TPM)",
  annotation_col = annotation_col,
  cluster_rows = TRUE,      
  cluster_cols = TRUE,    
  show_colnames = FALSE,    
  fontsize_row = 10
)
```
There's samples with no contamination at all but also samples across different organs contaminated with different markes. For this reason the most reasonable way to handle this complexity seems to add the average contamination score as a covariate. 

## 5. PCA

```{r}
pca = prcomp(log1p(t(tpm_matrix_filtered)), center = TRUE, scale. = TRUE)
eigs = pca$sdev^2
plot(1:length(eigs), eigs)
```
PCA plot colored by contamination score to detect a gradient (meaning that contamination contributes to variance)
```{r}
pca_df <- as.data.frame(pca$x) %>%
  rownames_to_column("Run") %>%
  left_join(meta_data, by = "Run")
ggplot(pca_df, aes(x = PC1, y = PC2, color = ContaminationScore, shape = Organ)) +
  geom_point(size = 4) +
  scale_color_viridis_c() + 
  labs(title = "PCA and impact of immune contamination",
       subtitle = "Samples with high contamination scores contribute to PC variance.") +
  theme_bw()
```

## 6. Pearson and Spearman Correlations

As per tutorial, plotting the dendograms based on the pearson and spearman correlations of the gene transcripts per million in the
samples:

```{r}

#calculating the pearson and spearman correlations
corr_pearson <- cor(log1p(tpm_matrix_filtered))
corr_spearman <- cor(tpm_matrix_filtered,method = "spearman")
hcl_pearson <- hclust(as.dist(1-corr_pearson))
hcl_spearman <-hclust(as.dist(1-corr_spearman))

#plotting dendograms
layout(matrix(1:2,nrow=1))
plot(hcl_pearson)
plot(hcl_spearman)

```

But unfortunately it does not seem to be very informative, there is a slight clustering of samples with close numbers, belonging to the same
organ. For example SR11256295,..296 and ...297 and SRR11256347,..348 and ..349.

##8. Saving the csv files necessary for the next step of the analysis
```{r}
write.csv(counts_matrix_filtered, "final_counts_filtered.csv", row.names = TRUE)
write.csv(tpm_matrix_filtered, "final_tpm_filtered.csv", row.names = TRUE)
write.csv(meta_data, "final_metadata_with_score.csv", row.names = FALSE)
write.csv(gene_map, "mgi_to_genesymbol_map.csv", row.names = FALSE)


print("Phase 1 complete. Files saved:")
print("- final_counts_filtered.csv (Use this for DESeq2)")
print("- final_metadata_with_score.csv (Input for DESeq2 Design)")
print("- final_tpm_filtered.csv (For visualizations)")
print("- mgi_to_genesymbol_map.csv (The gene ID map)")
```