---
title: "Data Curation and Correction"
author: "Marco Fioriani, Sara Leka, Carlo Rossetto"
date: "2025-11-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(dplyr)
  library(biomaRt)
  library(sva)
  library(pheatmap) 
  library(ggplot2)
  library(gridExtra)
  library(limma)
  library(DESeq2)
})
```

## Introduction

This document covers Part 1 of our analysis: Data Curation, Quality
Control, and Correction.

### Goals: - Load and format our aggregated count and TPM matrices. 
- Create and load our metadata file. 
- Filter out non-expressed genes (as per Tutorial.pdf, Sec 3-3). 
- Check for hematopoietic (immune cell) contamination, which was recognized as a risk of the FACS-sorting sample prep in Krausgruber et al. (2020). 
- Computationally regress out immune contamination signatures using limma::removeBatchEffect, as described by Krausgruber et al. (2020). 
- Save all the final, clean matrices for the next parts of our analysis.

## 1. Loading and preparing data
```{r}
counts_data = read.csv("final_counts_matrix.csv")
tpm_data = read.csv("final_tpm_matrix.csv")
meta_data = read.csv("SraRunTable.csv")

prep_matrix = function(df) {
  df_prepped = df %>%
  column_to_rownames(var = "gene_id")
  return(df_prepped)
}

counts_matrix = prep_matrix(counts_data)
tpm_matrix = prep_matrix(tpm_data)
```

## 2. Filtering out lowly-expressed genes
```{r}
avg_expr = rowMeans(tpm_matrix)
layout(matrix(1:2, nrow=1))
hist(avg_expr, main = "Histogram of Average TPM", xlab = "Average TPM")
hist(log10(avg_expr+ 1), main = "Histogram of Log10(TPM+1)", xlab = "Log10(TPM+1)")
```
```{r}
ggplot(data.frame(avg_expr), aes(x=avg_expr)) +
  geom_histogram(bins = 50) +
  scale_x_continuous(breaks = c(0,1,10,100,1000,10000,20000), trans="log1p",
expand=c(0,0)) +
  scale_y_continuous(breaks = c(0, 10, 100, 1000, 10000), trans="log1p") +
  theme_minimal() +
  ggtitle("Distribution of Average Gene Expression (Log scale)")
```

```{r}
num_det <- rowSums(tpm_matrix > 0)
hist(num_det, main = "Histogram of Gene Detections", xlab = "Number of Samples Gene is Detected In")
```
Filters for genes expressed in at least 50% of the samples or with an average expression of at least 1 TPM (saves genes that might be rare so detected in < 50% of samples but are highly expressed when they are present, e.g. a marker gene specific to just one organ)
```{r}
expressed = rowMeans(tpm_matrix > 0) >= 0.5 | rowMeans(tpm_matrix) >= 1
print(paste("Original gene count:", nrow(tpm_matrix)))
tpm_matrix_filtered = tpm_matrix[expressed, ]
counts_matrix_filtered = counts_matrix[expressed, ]

print(paste("New filtered gene count:", nrow(tpm_matrix_filtered)))
```

## 3. FACS contamination check by detection of immune cellmarkers

3a. Defining a contamination list based on the Krausgruber methods and the Han et al. (2018) "Mapping the Mouse Cell Atlas by Microwell-Seq". Using biomaRt to switch from MGI gene_id symbols to common gene symbols
```{r}
contamination_panel = list(
  "T_Cells" = c("Cd3d", "Cd3e", "Cd3g", "Lat", "Lck", "Vav1", "Tbx21", "Zap70"),
  "B_Cells" = c("Cd19", "Cd79a", "Cd79b", "Ms4a1"),
  "Macrophages" = c("Adgre1", "Cd68", "Csf1r"),
  "Monocytes" = c("Ly6c1", "Ly6c2", "Ccr2"),
  "NK_Cells" = c("Nkg7", "Klrb1c", "Gzma"),
  "Dendritic" = c("Itgax", "H2-Aa"),
  "Neutrophils" = c("Ly6g", "Cxcr2"),
  "Pan_Immune" = c("Ptprc")
)
immune_markers = unlist(contamination_panel)
mgi_gene_ids = rownames(tpm_matrix_filtered)

# connecting to gene mouse database
ensembl = useMart("ensembl", dataset = "mmusculus_gene_ensembl")

#creating a translation map 
gene_map <- getBM(
  attributes = c("mgi_symbol", "external_gene_name"), 
  filters = "mgi_symbol",
  values = mgi_gene_ids,         
  mart = ensembl
)
```

3b. Bar plotting the contamination per organ per immune cell type
```{r}
signature_scores = data.frame(Run = colnames(tpm_matrix_filtered))

for (cell_type in names(contamination_panel)) {
  genes = contamination_panel[[cell_type]]
  
  ids = gene_map$mgi_symbol[gene_map$external_gene_name %in% genes]
  ids = ids[ids %in% rownames(tpm_matrix_filtered)]
  
  if(length(ids) > 0) {
    gene_data = tpm_matrix_filtered[ids, , drop=FALSE]
    signature_scores[[cell_type]] = colMeans(log1p(gene_data))
  }
}

plot_data = signature_scores %>%
  left_join(meta_data %>% select(Run, Organ), by = "Run") %>%
  pivot_longer(cols = -c(Run, Organ), names_to = "CellType", values_to = "Score")

ggplot(plot_data, aes(x = Organ, y = Score, fill = Organ)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.2, size = 0.8, alpha = 0.6) + 
  facet_wrap(~ CellType, scales = "fixed", ncol = 4) + 
  theme_bw() +
  labs(
    title = "Contamination Burden by Cell Type",
    y = "Signature Score (Mean Log1p TPM)",
    x = ""
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    strip.background = element_rect(fill = "lightgrey")
  )
```
3c. Visualizing as heatmap to have more resolution on samples and genes (using VST, Variance Stabilizing Transformation, for visualization)
```{r}
heatmap_ids = gene_map$mgi_symbol[gene_map$external_gene_name %in% immune_markers]
heatmap_ids = heatmap_ids[heatmap_ids %in% rownames(counts_matrix_filtered)]

dds_temp = DESeqDataSetFromMatrix( 
  countData = round(counts_matrix_filtered), #VST on raw counts
  colData = meta_data,
  design = ~ Organ #so that it's not blind to organ difference, preserves biology
) 
vst_data = assay(vst(dds_temp, blind = FALSE)) 
heatmap_data_vst = vst_data[heatmap_ids, ]

annot_col = meta_data %>% select(Run, Organ) %>% column_to_rownames("Run")

gene_categories = data.frame(GeneID = character(), Type = character(), stringsAsFactors = FALSE)
for (type in names(contamination_panel)) {
  genes = contamination_panel[[type]]
  ids = gene_map$mgi_symbol[gene_map$external_gene_name %in% genes]
  ids = ids[ids %in% rownames(heatmap_data_vst)]
  if(length(ids) > 0) {
    gene_categories = rbind(gene_categories, data.frame(GeneID = ids, Type = type))
  }
}
annot_row = gene_categories %>% column_to_rownames("GeneID")

pheatmap(
  heatmap_data_vst,
  annotation_col = annot_col,
  annotation_row = annot_row,
  scale = "none",         
  breaks = seq(2, 14, length.out = 100),
  show_rownames = FALSE,
  show_colnames = FALSE,      
  main = "Comprehensive Contamination Check (VST Transformed)",
  cluster_rows = FALSE,        
  cluster_cols = TRUE,         
  fontsize_row = 8
)
```

##4. Handling FACS contamination (it will be used as a covariate in the DE design)

4a.Adding the contamination scores per contaminating immune cell type per sample to the metadata
```{r}
for (cell_type in names(contamination_panel)) {
  genes = contamination_panel[[cell_type]]
  
  ids = gene_map$mgi_symbol[gene_map$external_gene_name %in% genes]
  ids = ids[ids %in% rownames(tpm_matrix_filtered)]
  
  if(length(ids) > 0) {
    gene_data = tpm_matrix_filtered[ids, , drop=FALSE]
    score_name = paste0("Score_", cell_type)
    signature_scores[[score_name]] = colMeans(log1p(gene_data))
  }
}
meta_data = meta_data %>% left_join(signature_scores, by = "Run")
print(colnames(meta_data)) 
```
4b. Correction with limma for PCA and heatmaps
```{r}
score_cols = c("Score_T_Cells", "Score_Macrophages", "Score_Monocytes", "Score_Dendritic", "Score_NK_Cells", "Score_Pan_Immune", "Score_B_Cells")

covariate_matrix = as.matrix(meta_data[, score_cols])
rownames(covariate_matrix) = meta_data$Run

log_tpm = log1p(tpm_matrix_filtered)
design_matrix = model.matrix(~Organ, data = meta_data)

# Multi-variate correction:
# limma will  calculate a separate "slope" for Macrophages, NKs, etc.and subtract them independently from the relevant samples.
cleaned_log_tpm = removeBatchEffect(
  x = log_tpm,
  covariates = covariate_matrix,
  design = design_matrix
)
tpm_corrected = expm1(cleaned_log_tpm)
tpm_corrected[tpm_corrected < 0] = 0
```

4c. Visualize after limma::removeBatchEffetc multi-variate correction
```{r}
counts_corrected_rounded = round(counts_corrected)

dds_cor = DESeqDataSetFromMatrix(
  countData = counts_corrected_rounded, 
  colData = meta_data, 
  design = ~ Organ
)
vst_cor_data = assay(vst(dds_cor, blind = FALSE))

heatmap_ids = gene_map$mgi_symbol[gene_map$external_gene_name %in% immune_markers]
heatmap_ids = heatmap_ids[heatmap_ids %in% rownames(vst_cor_data)]

heatmap_data_clean = vst_cor_data[heatmap_ids, ]

annot_col = meta_data %>% select(Run, Organ) %>% column_to_rownames("Run")
gene_categories = data.frame(GeneID = character(), Type = character(), stringsAsFactors = FALSE)

for (type in names(contamination_panel)) {
  genes = contamination_panel[[type]]
  ids = gene_map$mgi_symbol[gene_map$external_gene_name %in% genes]
  ids = ids[ids %in% rownames(heatmap_data_clean)] 
  if(length(ids) > 0) {
    gene_categories = rbind(gene_categories, data.frame(GeneID = ids, Type = type))
  }
}
annot_row = gene_categories %>% column_to_rownames("GeneID")

pheatmap(
  heatmap_data_clean,
  annotation_col = annot_col,
  annotation_row = annot_row,
  scale = "none",
  breaks = seq(2, 14, length.out = 100),
  show_rownames = FALSE,       
  show_colnames = FALSE,       
  main = "Post-Correction Contamination Check (Multi-Variate VST)",
  cluster_rows = FALSE,        
  cluster_cols = TRUE,         
  fontsize_row = 8
)
```

4d. Compute composite score to correct for in downstream DGEX analysis
```{r}
meta_data$CompositeScore = rowMeans(meta_data[, score_cols])
summary(meta_data$CompositeScore)
```

## 5. PCA

5a. Helper function
```{r}
plot_pca_vst = function(count_matrix, title) {
  
  dds_temp = DESeqDataSetFromMatrix(
    countData = round(count_matrix),
    colData = meta_data,
    design = ~ Organ
  )
  
  vst_data = vst(dds_temp, blind = FALSE)
  vst_matrix = assay(vst_data)
  
  pca = prcomp(t(vst_matrix), center = TRUE, scale. = TRUE)
  
  pca_df = as.data.frame(pca$x) %>%
    rownames_to_column("Run") %>%
    left_join(meta_data, by = "Run")
  
  eigs = pca$sdev^2
  var_exp = round(eigs / sum(eigs) * 100, 1)
  
  ggplot(pca_df, aes(x = PC1, y = PC2, color = Organ, shape = Organ)) +
    geom_point(size = 7, alpha = 0.7) +
    scale_shape_manual(values = rep(c(16, 17, 15, 18), length.out = length(unique(pca_df$Organ)))) +
    theme_bw() +
    labs(
      title = title,
      x = paste0("PC1 (", var_exp[1], "%)"),
      y = paste0("PC2 (", var_exp[2], "%)")
    )
}
```

5b. Plotting pre and post
```{r, fig.width=14, fig.height=7, out.width="100%"}
#a. pre-correction data
p_vst_pre = plot_pca_vst(counts_matrix_filtered, "Before: Raw Counts (VST)")

#b. post-correction data
counts_corrected = expm1(cleaned_log_counts)
counts_corrected[counts_corrected < 0] = 0
counts_corrected = round(counts_corrected) 
p_vst_post = plot_pca_vst(counts_corrected, "After: Multi-Variate Correction (VST)")

grid.arrange(p_vst_pre, p_vst_post, ncol = 2)
```


## 6. Pearson and Spearman Correlations

As per tutorial, plotting the dendograms based on the pearson and spearman correlations of the gene transcripts per million in the
samples:

```{r}
#calculating the pearson and spearman correlations
corr_pearson <- cor(log1p(tpm_matrix_filtered))
corr_spearman <- cor(tpm_matrix_filtered,method = "spearman")
hcl_pearson <- hclust(as.dist(1-corr_pearson))
hcl_spearman <-hclust(as.dist(1-corr_spearman))

#plotting dendograms
layout(matrix(1:2,nrow=1))
plot(hcl_pearson)
plot(hcl_spearman)
```

But unfortunately it does not seem to be very informative, there is a slight clustering of samples with close numbers, belonging to the same
organ. For example SR11256295,..296 and ...297 and SRR11256347,..348 and ..349.

##7. Saving the csv files necessary for the next step of the analysis
```{r}
write.csv(counts_matrix_filtered, "final_counts_filtered.csv", row.names = TRUE)
write.csv(tpm_matrix_filtered, "final_tpm_filtered.csv", row.names = TRUE)
write.csv(tpm_corrected, "final_tpm_corrected.csv", row.names = TRUE) #mathematically corrected with multi-variate limma (for visualization)
write.csv(meta_data, "final_metadata_with_score.csv", row.names = FALSE)
write.csv(gene_map, "mgi_to_genesymbol_map.csv", row.names = FALSE)

print("Phase 1 complete. Files saved:")
print("- final_counts_filtered.csv (Use this for DESeq2)")
print("- final_metadata_with_score.csv (Input for DESeq2 Design)")
print("- final_tpm_filtered.csv (For visualizations)")
print("- final_tpm_corrected.csv (Clean matrix for Network Analysis/Figures)")
print("- mgi_to_genesymbol_map.csv (The gene ID map)")
```

```{r}
sessionInfo()
```

