---
title: "Data Curation and Correction"
author: "Marco Fioriani, Sara Leka, Carlo Rossetto"
date: "2025-11-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(dplyr)
  library(biomaRt)
  library(sva)
  library(pheatmap) 
  library(ggplot2)
})
```

##Introduction

This document covers Part 1 of our analysis: Data Curation, Quality Control, and Correction.

###Goals:
- Load and format our aggregated count and TPM matrices.
- Create and load our metadata file.
- Filter out non-expressed genes (as per Tutorial.pdf, Sec 3-3).
- Check for hematopoietic (immune cell) contamination, which is a known risk of the FACS-sorting sample prep (as per Krausgruber et al. 2020).
- Use ComBat_seq to correct for batch/individual effects, which also helps mitigate the contamination (as per Tutorial.pdf, Sec 3-3).
- Save the final, clean matrices for the next parts of our analysis.

##1. Loading and preparing data
```{r}
counts_data = read.csv("final_counts_matrix.csv")
tpm_data = read.csv("final_tpm_matrix.csv")
meta_data = read.csv("SraRunTable.csv")

prep_matrix = function(df) {
  df_prepped = df %>%
  column_to_rownames(var = "gene_id")
  return(df_prepped)
}

counts_matrix = prep_matrix(counts_data)
tpm_matrix = prep_matrix(tpm_data)
```

##2. Filter lowly-expressed genes
```{r}
avg_expr = rowMeans(tpm_matrix)
layout(matrix(1:2, nrow=1))
hist(avg_expr, main = "Histogram of Average TPM", xlab = "Average TPM")
hist(log10(avg_expr+ 1), main = "Histogram of Log10(TPM+1)", xlab = "Log10(TPM+1)")
```

```{r}
ggplot(data.frame(avg_expr), aes(x=avg_expr)) +
  geom_histogram(bins = 50) +
  scale_x_continuous(breaks = c(0,1,10,100,1000,10000,20000), trans="log1p",
expand=c(0,0)) +
  scale_y_continuous(breaks = c(0, 10, 100, 1000, 10000), trans="log1p") +
  theme_minimal() +
  ggtitle("Distribution of Average Gene Expression (Log scale)")
```

```{r}
num_det <- rowSums(tpm_matrix > 0)
hist(num_det, main = "Histogram of Gene Detections", xlab = "Number of Samples Gene is Detected In")
```

```{r}
expressed = rowMeans(tpm_matrix > 0) >= 0.5 | rowMeans(tpm_matrix) >= 1
print(paste("Original gene count:", nrow(tpm_matrix)))

tpm_matrix_filtered = tpm_matrix[expressed, ]
counts_matrix_filtered = counts_matrix[expressed, ]

print(paste("New filtered gene count:", nrow(tpm_matrix_filtered)))
```

##3. FACS contamination check (by detectionb of the pan-immune cell marker Ptprc)

3a. Using biomaRt to switch from MGI gene_id symbols to common gene symbols
```{r}
# connecting to gene mouse database
mgi_gene_ids = rownames(tpm_matrix_filtered)
ensembl <- useMart("ensembl", dataset = "mmusculus_gene_ensembl")

#creating a translation map 
gene_map <- getBM(
  attributes = c("mgi_symbol", "external_gene_name"), 
  filters = "mgi_symbol",
  values = mgi_gene_ids,         
  mart = ensembl
)

#finding the ptprc gene in our mgi gene_id
ptprc = gene_map$mgi_symbol[gene_map$external_gene_name == "Ptprc"]




# Check 1: Did we find the Ptprc gene in the biomaRt map?
if (length(ptprc) > 0 && !is.na(ptprc)) {
  
  # Check 2: Does that gene ID *also* exist in our *filtered* matrix?
  # (It could have been filtered out for low expression)
  if (ptprc %in% rownames(tpm_matrix_filtered)) {
    
    # 5. Extract Ptprc expression and join with metadata
    ptprc_expression <- tpm_matrix_filtered[ptprc, ] %>%
      t() %>%
      as.data.frame() %>%
      rownames_to_column(var = "Run") %>%
      left_join(meta_data, by = "Run")
    
    colnames(ptprc_expression)[2] <- "Ptprc_TPM"

    # 6. Plot the contamination
    print("Plotting expression of Ptprc (CD45) as a contamination check:")
    
    print( # We explicitly print the ggplot object inside the 'if' block
      ggplot(ptprc_expression, aes(x = Organ, y = Ptprc_TPM, fill = Organ)) +
        geom_boxplot() +
        geom_jitter(width = 0.1) +
        labs(
          title = "FACS Contamination Check: Ptprc (CD45) Expression",
          subtitle = "Ptprc is an immune cell marker and should be low in pure fibroblast samples.",
          y = "Ptprc Expression (TPM)",
          x = "Organ"
        ) +
        theme_bw() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
    )
    
  } else {
    print(paste("NOTE: 'Ptprc' (ID:", ptprc, ") was found in biomaRt but was FILTERED OUT for low expression."))
  }
  
} else {
  print("NOTE: 'Ptprc' gene not found after mapping with biomaRt.")
  print("This is OK, but it means we can't use it as our only QC check.")
  print("We will proceed with ComBat_seq correction as planned.")
}
```
```{r}
# --- ADD THIS CODE to the end of your "check-contamination" chunk ---

contamination_table <- ptprc_expression %>%
  dplyr::filter(Organ %in% c("Kidney", "Spleen")) %>%
  dplyr::select(Run, Organ, BioSample, Ptprc_TPM) %>%
  dplyr::arrange(desc(Ptprc_TPM)) # Sort by highest contamination

knitr::kable(contamination_table, caption = "Ptprc (CD45) Contamination per Sample")
```
##4. Handling FACS contamination
###...

##5. PCA
```{r}
pca = prcomp(log1p(t(tpm_matrix_filtered)), center = TRUE, scale. = TRUE)
eigs = pca$sdev^2
plot(1:length(eigs), eigs)
```

```{r}
ggplot(data.frame(pca$x, meta_data)) +
  geom_point(aes(x = PC1, y = PC2, color = Organ), size = 5)
```

```{r}
setwd("~/My Documents/Carlo/ETH UBas UZH - MSc CBB/636-0101-00L Systems Genomics/02 Tutorial (RNA Sequencing Project)/05_GitHub_R_Work/RNAseq_drylab")

write.csv(counts_matrix_filtered, "final_counts_filtered.csv", row.names = TRUE)

write.csv(tpm_matrix_filtered, "final_tpm_filtered.csv", row.names = TRUE)

write.csv(gene_map, "mgi_to_genesymbol_map.csv", row.names = FALSE)


print("Phase 1 complete. Files saved:")
print("- final_counts_filtered.csv (Use this for DESeq2)")
print("- final_tpm_filtered.csv (For visualizations)")
print("- mgi_to_genesymbol_map.csv (The gene ID map)")
```

##6. Visualisations

As per tutorial, plotting the dendograms based on the pearson and spearman correlations of the gene transcripts per million in the samples:
```{r}

#calculating the pearson and spearman correlations
corr_pearson <- cor(log1p(tpm_matrix_filtered))
corr_spearman <- cor(tpm_matrix_filtered,method = "spearman")
hcl_pearson <- hclust(as.dist(1-corr_pearson))
hcl_spearman <-hclust(as.dist(1-corr_spearman))

#plotting dendograms
layout(matrix(1:2,nrow=1))
plot(hcl_pearson)
plot(hcl_spearman)

```
But unfortunately it does not seem to be very informative, there is a slight clustering of samples with close numbers, belonging to the same organ. For example SR11256295,..296 and ...297 and SRR11256347,..348 and ..349. 

##7. Differential Expression Analysis

###Approach 1: ANOVA and ANCOVA, our own method
```{r}
# to be attempted
```


###Approach 2: DESeq2

#### Calculating the Log2fold changes
```{r}
library(DESeq2)

# preparing the tables (using the counts_matrix_filtered as recommended)
# converting the counts to integers (by rounding them) and the table type to matrix
rounded_cmf = round(counts_matrix_filtered,digits = 0)
rounded_cmf <- as.matrix(rounded_cmf)
storage.mode(rounded_cmf) <- "integer"

# a table with two columns, sample names and organs should is built
col_data = meta_data[c("Run","Organ")]
col_data$Organ <- factor(col_data$Organ)
levels(col_data$Organ)

# the levels of the organs (not in the table) are rearranged, the reference organ is the one other organ expression levels will be compared against
col_data$Organ <- relevel(col_data$Organ, ref = "Liver")
levels(col_data$Organ) 


# generating the differential expression data
dds_filtered <-  DESeqDataSetFromMatrix(countData = rounded_cmf, colData = col_data,design = ~ Organ)
dds_filtered <- DESeq(dds_filtered, test="LRT", reduced = ~1)
res_DESeq2 <- results(dds_filtered)
```
Visualising the DESeq 2 results

####Visualising

This first plot is just the log2fold change and normalised gene count for two organs against the reference.
```{r}
res <- results(dds_filtered, contrast = c("Organ", "SmallIntestine","LargeIntestine"))
plotMA(res, main="MA-plot:SmallIntestine and Liver vs Kidney", ylim=c(-10,10))
legend("topright", legend=c("small intestine", "liver"), 
       col=c("blue", "grey"), pch=16, bty="n")
```

Visualising the normalised counts of the top 20 genes
```{r}
library(pheatmap)
vsd <- vst(dds_filtered, blind=FALSE)  # variance stabilizing transform
top_genes <- head(order(res$padj), 20)  # top 20 DE genes
norm_counts <- assay(vsd)[top_genes, ]
pheatmap(norm_counts, cluster_rows=TRUE, cluster_cols=TRUE,
         annotation_col=as.data.frame(colData(dds_filtered)["Organ"]),
         main="Normalised counts heatmap of top 20 most differentially expressed genes") # note this plot is independent of the reference organ chosen, because it is the normalised count, and the top 20 differentially expressed genes remain the same independently of the reference gene organ chosen (?)
```

more DESeq2 plotting, now the log fold changes
```{r}
organ_levels <- levels(col_data$Organ)
organ_levels <- organ_levels[organ_levels != "Liver"]  # exclude reference (the same organ that was set as reference three chunks ago)

# Create a matrix to hold log2 fold changes: genes x organs
log2fc_matrix <- sapply(organ_levels, function(org) {
  res_org <- results(dds_filtered, contrast=c("Organ", org, "Liver"))
  log2fc <- res_org$log2FoldChange
  names(log2fc) <- rownames(res_org)
  return(log2fc)
})

# Convert to matrix and remove genes with NA values in any organ
log2fc_matrix <- as.matrix(log2fc_matrix)
log2fc_matrix <- log2fc_matrix[complete.cases(log2fc_matrix), ]

# Generate heatmap of log2 fold changes
pheatmap(log2fc_matrix,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames = FALSE,
         main = "Heatmap of log2 fold changes across organs compared to liver",
         fontsize_col = 12,
         color = colorRampPalette(c("blue", "white", "red"))(100))
```

same as last plot but only with top genes:
```{r}
# Calculate variance for each gene (row)
gene_variances <- apply(log2fc_matrix, 1, var)

# Select top N genes by variance, for example top 20
top_genes <- order(gene_variances, decreasing = TRUE)[1:20]

# Subset the log2fc matrix to include only top variable genes
log2fc_top <- log2fc_matrix[top_genes, ]

# Now you can use 'log2fc_top' to plot the heatmap
library(pheatmap)
pheatmap(log2fc_top,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames = TRUE,
         main = "Log2Fold Change of Top 20 Most Differentially Expressed Genes Compared to Liver",
         legend=TRUE,
         annotation_legend=TRUE,
         color = colorRampPalette(c("blue", "white", "red"))(100))
```

Comparing with the ANOVA and ANCOVA gene expression: approach 1 vs approach 2
```{r}
#to be continued
#cor(res_DESeq2$padj, res_DE %>% filter(gene %in% rownames(res_DESeq2)) #%>% pull(padj), method = "spearman", use="complete.obs")

```