---
title: "3_differential_gex_ANOVA"
author: "Marco Fiorani, Sara Leka and Carlo Rossetto"
date: "2025-11-30"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(dplyr)
  library(biomaRt)
  library(sva)
  library(pheatmap) 
  library(ggplot2)
  library(gridExtra)
})
```

For differential expression, three approaches were used:  Limma (as per paper), ANOVA (as per tutorial) and DESeq2 (as per tutorial) and. In this script the second approach is shown.

### Approach 2: ANOVA, our own method as per tutorial

```{r}
DE_test <- function(expr, cond, ctrl = NULL, covar = NULL, padj_method = p.adjust.methods) {
  
  # Use proper data frame construction instead of t()
  results <- pbapply(expr, 1, function(e) {
    
    dat <- data.frame(y = log1p(e), cond = cond)
    if (!is.null(covar)) dat <- data.frame(dat, covar)
    
    m1 <- lm(y ~ ., data = dat)
    m0 <- lm(y ~ . - cond, data = dat)
    test <- anova(m1, m0)
    pval <- test$Pr[2]
    
    # Calculate mean log-expression per condition
    avgs <- tapply(log1p(e), cond, mean)
    
    # Find organ with maximum mean expression
    max_organ_idx <- which.max(avgs)
    max_organ <- names(avgs)[max_organ_idx]
    max_organ_mean <- avgs[max_organ_idx]
    
    # Fold change: max organ vs mean of ALL organs
    all_organs_mean <- mean(avgs)
    fc <- exp(max_organ_mean - all_organs_mean)
    
    # Return as data frame row (handles mixed types)
    data.frame(
      pval = as.numeric(pval),
      fc = as.numeric(fc),
      max_organ = as.character(max_organ),
      stringsAsFactors = FALSE
    )
  })
  
  # Combine into single data frame
  pval_fc_maxorgan <- do.call(rbind, results)
  rownames(pval_fc_maxorgan) <- rownames(expr)
  
  # Multiple-testing correction
  padj <- p.adjust(pval_fc_maxorgan$pval, method = padj_method)
  
  # Return result table
  return(data.frame(
    pval_fc_maxorgan, 
    padj = padj
  )[, c("pval", "padj", "fc", "max_organ")])
}

```


```{r}
suppressPackageStartupMessages({library(pbapply)}) 

res_DE <- DE_test(
  expr  = counts_matrix_filtered,
  cond  = meta_data$Organ,
  covar = NULL
) %>% tibble::rownames_to_column("gene")
```

```{r}
suppressPackageStartupMessages({library(pbapply)}) 

res_DE_contamination <- DE_test(
  expr  = counts_matrix_filtered,
  cond  = meta_data$Organ,
  covar = meta_data$CompositeScore
)  %>% tibble::rownames_to_column("gene")
```


```{r}
# 1. Create DE flags (MANDATORY) - NOW for BOTH datasets

res_DE <- res_DE %>%
  mutate(DE  = padj < 0.05 & fc > 2) %>%
  mutate(DEG = ifelse(DE, .data$gene, NA_character_))

res_DE_contamination <- res_DE_contamination %>%
  mutate(DE  = padj < 0.05 & fc > 2) %>%
  mutate(DEG = ifelse(DE, .data$gene, NA_character_))

```

```{r}
suppressPackageStartupMessages({library(ggrepel)})

ggplot(res_DE, aes(x = log(fc), y = -log10(padj), col=DE, label=DEG)) +
  geom_point() +
  geom_text_repel() +
  geom_vline(xintercept=c(log(2), 0), col="#303030", linetype="dotted") +
  geom_hline(yintercept=-log10(0.05), col="#303030", linetype="dotted") +  # Fixed: 0.05
  scale_color_manual(values=c("FALSE"="#909090", "TRUE"="red")) +
  theme_minimal()
```

```{r}
library(ggplot2)
library(dplyr)
library(ggpubr)

# Combine results
res_combined <- bind_rows(
  res_DE %>% mutate(method = "No Covariate"),
  res_DE_contamination %>% mutate(method = "With Contam. Covar"),
)

# 1. Volcano plot overlay
p1 <- ggplot(res_combined, aes(x = log(fc), y = -log10(padj), color = method)) +
  geom_point(alpha = 0.6, size = 1) +
  facet_wrap(~method) +
  labs(title = "Volcano Plot Comparison") +
  theme_bw()

print(p1)
```

```{r}
# 2. P-value difference plot
res_diff <- inner_join(
  res_DE %>% dplyr::select(gene, pval_no_covar = pval, fc_no_covar = fc, DE),
  res_DE_contamination %>% dplyr::select(gene, pval_contam = pval, fc_contam = fc),
  by = "gene"
) %>%
  mutate(pval_diff = pval_no_covar - pval_contam)

p2 <- ggplot(res_diff, aes(x = fc_no_covar, y = pval_diff, color = DE)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_x_log10() +
  scale_color_manual(values = c("FALSE" = "grey50", "TRUE" = "red"), 
                     name = "DE (padj<0.05,\nFC>2)", 
                     labels = c("No", "Yes")) +
  labs(x = "FC (No Covar, log10)", y = "Δp-value (No Covar - Contam)", 
  title = "P-value shift by contamination correction") +
  theme_bw()

# same plot only with only the differentially expressed genes and the ones that lose their significance after contamination

# Filter to ONLY DE genes for the second plot
res_diff_DE <- res_diff %>%
  filter(DE == TRUE)  # Keep only DE genes (padj < 0.05 & fc > 2)
```

```{r}
# EXPLICIT: Only genes that were DE in res_DE (no covariate)
res_diff_DE_only <- inner_join(
  res_DE %>% 
    filter(DE == TRUE) %>%  # ONLY DE genes from res_DE
    dplyr::select(gene, pval_no_covar = pval, fc_no_covar = fc, padj_no_covar = padj),
  res_DE_contamination %>% 
    dplyr::select(gene, pval_contam = pval, fc_contam = fc, padj_contam = padj),
  by = "gene"
) %>%
  mutate(
    pval_diff = pval_no_covar - pval_contam,
    lost_sig = padj_no_covar < 0.05 & padj_contam >= 0.05
  )

# Plot: ONLY DE genes from res_DE
p2_DE_only <- ggplot(res_diff_DE_only, aes(x = fc_no_covar, y = pval_diff, color = lost_sig)) +
  geom_point(alpha = 0.6, size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  scale_x_log10() +
  scale_color_manual(
    values = c("FALSE" = "red", "TRUE" = "orange"), 
    name = "Lost sig\nafter contam",
    labels = c("Remains DE", "Lost DE")
  ) +
  labs(x = "FC (No Covar, log10)", y = "Δp-value", 
       title = "ONLY genes DE in res_DE (no covariate)") +
  theme_bw()

print(p2)
print(p2_DE_only)
```

```{r}
# 3. Bland-Altman for padj (SIMPLE approach)
res_padj <- inner_join(
  res_DE %>% 
    dplyr::select(gene, padj_no_covar = padj, DE_no_covar = DE),
  res_DE_contamination %>% 
    dplyr::select(gene, padj_contam = padj, DE_contam = DE),
  by = "gene"
) %>%
  mutate(
    mean_padj = (padj_no_covar + padj_contam)/2,
    diff_padj = padj_no_covar - padj_contam,
    # 3-color logic:
    color_group = case_when(
      DE_contam == TRUE ~ "1: DE with contam covar",                    # Green: sig with contamination
      DE_no_covar == TRUE & DE_contam == FALSE ~ "2: Lost sig",         # Orange: sig→NOT sig
      TRUE ~ "3: Remaining genes"                                       # Grey: everything else
    )
  ) %>%
  mutate(color_group = factor(color_group, levels = c("1: DE with contam covar", 
                                                      "2: Lost sig", 
                                                      "3: Remaining genes")))

p3 <- ggplot(res_padj, aes(x = padj_no_covar, y = diff_padj, color = color_group)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(alpha = 0.6, size = 1) +
  scale_x_log10(limits = c(1e-4, 1)) +  # Ensure all genes plotted
  scale_color_manual(
    values = c("1: DE with contam covar" = "green4", 
               "2: Lost sig" = "orange", 
               "3: Remaining genes" = "grey60"),
    name = "DE Status",
    labels = c("DE with contam", "Lost sig", "Remaining")
  ) +
  labs(x = "padj No Covar (log10)", y = "Δpadj (No Covar - Contam)", 
       title = "Bland-Altman: padj agreement (3 colors)") +
  theme_bw()

print(p3)
```

### 4. Summary table

```{r}
# Summary table of DEGs - FIXED
de_summary <- bind_rows(
  res_DE %>% mutate(method = "No Covariate"),
  res_DE_contamination %>% mutate(method = "With Contam. Covar")
) %>%
  summarise(
    Total_Genes = n(),
    DEGs = sum(DE, na.rm = TRUE),
    .by = method  # Group by method column
  ) %>%
  mutate(
    Prop_DE = scales::percent(DEGs / Total_Genes, accuracy = 0.1)
  ) %>%
  dplyr::select(Method = method, Total_Genes, DEGs, Prop_DE)

# Pretty table
knitr::kable(de_summary, 
             col.names = c("Model", "Total Genes", "DEGs<br>(padj<0.05, FC>2)", "% DE"),
             digits = 0)

```

``
The contamination seems significant!

```{r}
# Get genes that were DE before (res_DE) but NOT after contamination correction
lost_DE_genes <- inner_join(
  res_DE %>% dplyr::select(gene, padj_no_covar = padj),
  res_DE_contamination %>% dplyr::select(gene, padj_contam = padj),
  by = "gene"
) %>%
  filter(padj_no_covar < 0.05 & padj_contam >= 0.05) %>%  # Sig before, NOT after
  pull(gene)

# Output the list
print("Genes that lost significance after contamination correction:")
print(lost_DE_genes)

# Optional: Save to file
write.csv(data.frame(gene = lost_DE_genes), 
          "lost_DE_genes_contamination.csv", 
          row.names = FALSE)

# Count how many
cat("Total genes that lost DE status:", length(lost_DE_genes), "\n")
```
Adding the covariance does not influence the fold change 

Note the approach above finds the maximum and minimum count amongst organs and computes one fold change per gene. So one gene fold change may be: 
max gene count from sample in liver - max gene count from sample in brain, while another may be: max gene count from sample in kidney - max gene count from sample in small intestine, there is one "fold change" value per gene.

Also the fold change is linear, not log(base 2) (?).

```{r}
de_joined <- inner_join(res_DE, res_DE_contamination, by = "gene")

de_matrix <- de_joined %>%
  mutate(
    sig_no_covar  = padj.x  < 0.05 & fc.x  > 2,
    sig_contam    = padj.y    < 0.05 & fc.y    > 2,
    DE_status = case_when(
      sig_no_covar  & sig_contam  ~ "significant in both",
      sig_no_covar  & !sig_contam ~ "significant only without contamination",
      !sig_no_covar & sig_contam  ~ "significant only with contamination"
      # no "not significant in either" level
    )
  ) %>%
  # keep only rows that are significant in at least one model
  filter(!is.na(DE_status)) %>%
  dplyr::select(
    gene,
    padj  = padj.y,
    pval  = pval.y,
    fc    = fc.y,
    max_organ = max_organ.y,
    DE_status
  )

de_matrix_df = as.data.frame(de_matrix)

```




```{r}
de_by_organ_base <- table(de_matrix[,"max_organ"])
print(de_by_organ_base)

# 1. Total DEGs per organ (what you have)
de_by_organ_total <- as.data.frame(table(de_matrix[,"max_organ"]), 
                                   stringsAsFactors = FALSE)
colnames(de_by_organ_total) <- c("max_organ", "Total_DEGs")

# 2. DEGs without contamination (DE only without + both)
no_covar_idx <- grepl("without|both", de_matrix[,"DE_status"], ignore.case = TRUE)
de_no_covar <- table(de_matrix[no_covar_idx, "max_organ"])

# 3. DEGs with contamination (DE only with + both)  
with_covar_idx <- grepl("both|with ", de_matrix[,"DE_status"], ignore.case = TRUE)
de_with_covar <- table(de_matrix[with_covar_idx, "max_organ"])

# 4. Combine into final summary
summary_table <- data.frame(
  max_organ = names(de_no_covar),
  DEGs_no_covar = as.numeric(de_no_covar),
  DEGs_with_covar = as.numeric(de_with_covar),
  Total_DEGs = as.numeric(table(de_matrix[,"max_organ"])[names(de_no_covar)])
)

print(summary_table)
```


```{r}
# Use de_matrix_df (data frame version) - top 10 FC genes per organ
de_matrix_df$fc <- as.numeric(de_matrix_df$fc)

top10_genes <- de_matrix_df %>%
  group_by(max_organ) %>%
  slice_max(fc, n = 10) %>%  # Top 10 highest FC per organ
  ungroup() %>%
  dplyr::select(gene, max_organ, fc)

# Create heatmap matrix: genes x organs, fill with NA for non-max organs
heatmap_mat <- matrix(NA, nrow = nrow(top10_genes), ncol = length(unique(de_matrix_df$max_organ)))
rownames(heatmap_mat) <- top10_genes$gene
colnames(heatmap_mat) <- unique(top10_genes$max_organ)

# Fill only the max_organ column with FC values
for(i in 1:nrow(top10_genes)) {
  organ_col <- which(colnames(heatmap_mat) == top10_genes$max_organ[i])
  heatmap_mat[i, organ_col] <- top10_genes$fc[i]
}

# Plot heatmap (NA = white, high FC = red)
heatmap_mat_no_na <- heatmap_mat
heatmap_mat_no_na[is.na(heatmap_mat_no_na)] <- 0
pheatmap(heatmap_mat_no_na,
         scale = "none",
         color = colorRampPalette(c("white", "red"))(100),
         cluster_rows = TRUE,
         cluster_cols = FALSE,
         main = "Top 10 DEGs per organ (FC)",
         na_col = "grey90")  # Low expression = light grey
```


```{r}
# 1. No contamination DEGs per organ
no_covar_idx <- grepl("without|both", de_matrix[,"DE_status"], ignore.case = TRUE)
no_covar_genes <- de_matrix[no_covar_idx, c("gene", "max_organ")]

no_covar_lists <- split(no_covar_genes$gene, no_covar_genes$max_organ)
names(no_covar_lists) <- paste0(names(no_covar_lists), "_no_covar")

# 2. With contamination DEGs per organ  
with_covar_idx <- grepl("both|with ", de_matrix[,"DE_status"], ignore.case = TRUE)
with_covar_genes <- de_matrix[with_covar_idx, c("gene", "max_organ")]

with_covar_lists <- split(with_covar_genes$gene, with_covar_genes$max_organ)
names(with_covar_lists) <- paste0(names(with_covar_lists), "_with_covar")

# 3. All gene lists ready for GO enrichment
all_go_lists <- c(no_covar_lists, with_covar_lists)

# Quick check
print("No covar lists:")
lapply(no_covar_lists, length)
print("With covar lists:")
lapply(with_covar_lists, length)

# Now run GO on any list: enrichGO(gene = all_go_lists[["Liver_no_covar"]], ...)
```
