---
title: "Differential Gene Expression Analysis"
author: "Marco Fioriani, Sara Leka, Carlo Rossetto"
date: "2025-11-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(dplyr)
  library(biomaRt)
  library(sva)
  library(pheatmap) 
  library(ggplot2)
  library(gridExtra)
})
```


### Approach 1 - Limma package following Krausgruber et al.

```{r}
suppressPackageStartupMessages({library(pbapply)}) 

res_DE <- DE_test(
  expr  = counts_matrix_filtered,
  cond  = meta_data$Organ,
  covar = meta_data$ContaminationScore
) %>%
  tibble::rownames_to_column("gene")
```

```{r}
# 2. Create DE flags (MANDATORY)
res_DE <- res_DE %>%
  mutate(DE  = padj < 0.1 & fc > 2) %>%   # TRUE/FALSE
  mutate(DEG = ifelse(DE, gene, NA))      # label only DE genes

suppressPackageStartupMessages({library(ggrepel)})
ggplot(res_DE, aes(x = log(fc), y = -log10(padj), col=DE, label=DEG)) +
  geom_point() +
  geom_text_repel() +
  geom_vline(xintercept=c(log(2), 0), col="#303030", linetype="dotted") +
  geom_hline(yintercept=-log10(0.1), col="#303030", linetype="dotted") +
  scale_color_manual(values=c("#909090", "red")) +
  theme_minimal()
```

Note the approach above finds the maximum and minimum count amongst organs and computes one fold change per gene. So one gene fold change may be: 
max gene count from sample in liver - max gene count from sample in brain, while another may be: max gene count from sample in kidney - max gene count from sample in small intestine, there is one "fold change" value per gene.

Also the fold change is linear, not log(base 2) (?).

### Approach 2: DESeq2

#### Calculating the Log2fold changes

```{r}
suppressPackageStartupMessages(library(DESeq2))

# preparing the tables (using the counts_matrix_filtered as recommended)
# converting the counts to integers (by rounding them) and the table type to matrix
rounded_cmf = round(counts_matrix_filtered,digits = 0)
rounded_cmf <- as.matrix(rounded_cmf)
storage.mode(rounded_cmf) <- "integer"

# a table with two columns, sample names and organs should is built
col_data = meta_data[c("Run","Organ")]
col_data$Organ <- factor(col_data$Organ)
levels(col_data$Organ)

# the levels of the organs (not in the table) are rearranged, the reference organ is the one other organ expression levels will be compared against
col_data$Organ <- relevel(col_data$Organ, ref = "Liver")
levels(col_data$Organ) 


# generating the differential expression data
dds_filtered <-  DESeqDataSetFromMatrix(countData = rounded_cmf, colData = col_data,design = ~ Organ)
dds_filtered <- DESeq(dds_filtered, test="LRT", reduced = ~1)
res_DESeq2 <- results(dds_filtered)
```

The default results(dds_filtered) after an LRT gives you this global
p‑value per gene, plus one particular log2FC column (often the first
non‑reference level vs reference), but the p‑value is for “any organ
difference,” not for that single pairwise contrast.

If you want specific organ vs organ log2FCs after fitting with LRT, you
still call results(dds_filtered, contrast = c("Organ", "Kidney",
"Liver")), which will use the already estimated coefficients to compute
that pairwise log2FC; the LRT p‑values remain the same global ones.

(1) key difference is that the fold changes, in comparison to the ANOVA
    above, are log base 2. (2) The log2fold changes are in comparison to
    one reference organ, and not max() - min() across all organs like in
    ANOVA. To use RE_Seq2 to calculate log2fold changes against a mean
    count of all organs one would either have to introduce an artificial
    factor/organ that is the calculated mean across all organs or
    calculate it manually by iterating through all organs and dividing
    by the averaged mean count for that gene.

Next is some visualisation of the DESeq 2 results, it is extra, not in
tutorial.

####Visualising 

This section is extra (not from tutorial), the section
"8. Grouping DEGs" has the same scope and follows the tutorial, so that one is better to follow.

This first plot is the log2fold change versus normalised gene count for two organs (average of samples across the organ) against the reference organ (its samples averaged). Really just for exploration.

```{r}
res <- results(dds_filtered, contrast = c("Organ", "SmallIntestine","LargeIntestine"))
plotMA(res, main="MA-plot:SmallIntestine and Liver vs Kidney", ylim=c(-10,10))
legend("topright", legend=c("Small intestine", "Large intestine"), 
       col=c("blue", "grey"), pch=16, bty="n")
```

Visualising the normalised counts of the top 20 differentially expressed genes based on DESeq2.

```{r}
library(pheatmap)
vsd <- vst(dds_filtered, blind=FALSE)  # variance stabilizing transform
top_genes <- head(order(res$padj), 20)  # top 20 differentially expressed genes considering small intestine and large intestine versus reference organ, liver
norm_counts <- assay(vsd)[top_genes, ]
pheatmap(norm_counts, cluster_rows=TRUE, cluster_cols=TRUE,
         annotation_col=as.data.frame(colData(dds_filtered)["Organ"]),
         main="Normalised counts heatmap of top 20 most differentially expressed genes")
```

more DESeq2 plotting, now the log base2 fold changes of all the
differentially expressed genes (according to DESeq2)

```{r}
organ_levels <- levels(col_data$Organ)
organ_levels <- organ_levels[organ_levels != "Liver"]  # exclude reference (the same organ that was set as reference three chunks ago)

# Create a matrix to hold log2 fold changes: genes x organs
log2fc_matrix <- sapply(organ_levels, function(org) {
  res_org <- results(dds_filtered, contrast=c("Organ", org, "Liver"))
  log2fc <- res_org$log2FoldChange
  names(log2fc) <- rownames(res_org)
  return(log2fc)
})

# Convert to matrix and remove genes with NA values in any organ
log2fc_matrix <- as.matrix(log2fc_matrix)
log2fc_matrix <- log2fc_matrix[complete.cases(log2fc_matrix), ]

# Generate heatmap of log2 fold changes
pheatmap(log2fc_matrix,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames = FALSE,
         main = "Heatmap of log2 fold changes across organs compared to liver",
         fontsize_col = 12,
         color = colorRampPalette(c("blue", "white", "red"))(100))
```

same as last plot but only with top 20 most differentially expressed
genes acrossed all organs (the ones with the highest variance in
log2fold change amongs all organs against the reference organ):

```{r}
# Calculate variance for each gene (row)
gene_variances <- apply(log2fc_matrix, 1, var)

# Select top N genes by variance, for example top 20
top_genes <- order(gene_variances, decreasing = TRUE)[1:20]

# Subset the log2fc matrix to include only top variable genes
log2fc_top <- log2fc_matrix[top_genes, ]

# Now you can use 'log2fc_top' to plot the heatmap
library(pheatmap)
pheatmap(log2fc_top,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames = TRUE,
         main = "Log2Fold Change of Top 20 Most Differentially Expressed Genes Compared to Liver",
         legend=TRUE,
         annotation_legend=TRUE,
         color = colorRampPalette(c("blue", "white", "red"))(100))
```

Comparing DESeq2 with the ANOVA and ANCOVA gene expression: approach 1
vs approach 2

```{r}

cor(res_DESeq2$padj, res_DE %>% filter(gene %in% rownames(res_DESeq2)) %>% pull(padj), method = "spearman", use="complete.obs")

layout(matrix(1:2, nrow = 1))

## standard scatter plot
plot(
  -log10(res_DESeq2$pvalue),
  -log10(
    res_DE %>%
      dplyr::filter(gene %in% rownames(res_DESeq2)) %>%
      dplyr::pull(pval)
  ),
  xlab = "-log10(pval DESeq2)",
  ylab = "-log10(pval DE)",
  pch  = 16
)

## smooth density scatter plot
smoothScatter(
  -log10(res_DESeq2$pvalue),
  -log10(
    res_DE %>%
      dplyr::filter(gene %in% rownames(res_DESeq2)) %>%
      dplyr::pull(pval)
  ),
  xlab = "-log10(pval DESeq2)",
  ylab = "-log10(pval DE)",
  pch  = 16
)

```

```{r}
table(
  p.adjust(res_DESeq2$pvalue, method = "bonferroni") < 0.1,
  res_DE %>%
    dplyr::filter(gene %in% rownames(res_DESeq2)) %>%
    dplyr::pull(padj) < 0.1
)
```

### Approach 3 - Limma package following Krausgruber et al.

In this section instead of using ANOVA or DESeq2 (file 3_differential_gex_ANOVA_DESeq2) the differentialexpression package recommended by the article is used: Limma. The
articles figures that were generated with the steady state fibroblast
RNA data are then replicated.

Fig.1d - Finding Differential Expressed genes of each organs vs all
Calculate log2(CPM). 1) z-scores between replicates and between organ
with heatmaps 2) Marker gene expressed genes are filtered with log2(FC)
\> 0.5 by limma function that use (Log2(CPM)) and with padj \< 0.1 At
the end you get "marker_genes_by_organ" that contains all the marker
genes upregulated for that organ (vs others) after the filter above
Output: lenght of significant (padj) marker gene found per organ

```{r}
suppressPackageStartupMessages({library(edgeR) #handles counts, CPM, normalization.
                                library(limma)
                                library(gplots)
                                library(viridis)}) #performs the linear modeling we will use for “organ vs others”

#This converts the counts_matrix into an RNA-seq object that edgeR/limma understand.
counts_matrix_limma <- DGEList(counts = counts_matrix_filtered)

#Create a matrix where it says for each sample to which organ it belongs, (just covert col_data in a "limma" compatible version)
design <- model.matrix(~ 0 + col_data$Organ) 

#Prepares the data for the limma differential expression: Converts counts → log2(CPM), Estimates precision weights per gene and 
voom_object <- voom(counts_matrix_limma, design, plot = FALSE)

# matrice genes × samples, in scala log2(CPM)
log2cpm_matrix_limma <- voom_object$E   

#z-scores per gene (for later heatmaps)
expr_z <- t(scale(t(log2cpm_matrix_limma)))


#STEP 1 Average expression of each gene within replicates with lmFit
organs <- sort(unique(col_data$Organ))     

#Create a design matrix dummy that has just 1/0 in rows. It indicates which sample belongs to which organ
design_org <- model.matrix(~ 0 + col_data$Organ)

#Change the name of colums in more readable organs
colnames(design_org) <- organs

#Calculates average expression in each organ and uses replicates to estimate variances and standard errors (Example: Gene X in Brain: mean = 8.5, SE = 0.3)
fit0 <- lmFit(log2cpm_matrix_limma, design_org)


#PRELIMINARY HEATMAP
heatmap(expr_z, col = colorRampPalette(c("blue", "white", "red"))(50))


mean_by_organ <- t(apply(log2cpm_matrix_limma, 1, 
                        function(x) tapply(x, col_data$Organ, mean)))

z_scores_organ <- t(scale(t(mean_by_organ)))

heatmap(z_scores_organ, col = colorRampPalette(c("blue", "white", "red"))(50))
```

```{r}
#STEP 2 Prepare all pairwise contrasts matrix (A vs B)
contrast_list <- c()

#For any organ
for (o in organs) {
  #takes all organs except "o" selected
  others <- setdiff(organs, o)
  #For any other organ
  for (b in others) {
    #Create name like "Brain_vs_Liver" for me(user)
    cname <- paste0(o, "_vs_", b)
    #Create the limma compatible name "Liver-Brain"
    contrast_list[cname] <- paste0(o, "-", b)
  }
}

#Preparing binary/dummy matrix of contrast to do. For each key "Brain_vs_Liver" takes the value "Liver-Brain" = (+1 × Liver) + (-1 × Brain) and 0 for other...
contrast_matrix <- makeContrasts(contrasts = contrast_list,
                                 levels    = design_org)


#TEP 3 – Get DE results for all organ pairs18

fit_all_contrast <- contrasts.fit(fit0, contrast_matrix) #Apply contrasts from the matrix fit0 --> contains logFC for each contrast
fit_all_contrast <- eBayes(fit_all_contrast) #Bayesian statistics adds: t, p.value, lods, F, F.p.value to each logFC constrast

pair_res_list <- list()

#for each contrast
for (o_vs_o_contrast in colnames(contrast_matrix)) {
  #extract the result column of that specif contrast "o_vs_o_contrast" from "fit_all_contrast" and puts in "tt" (toptable)
  tt <- topTable(fit_all_contrast, coef = o_vs_o_contrast, number = Inf) 
  tt$gene     <- rownames(tt)
  tt$contrast <- o_vs_o_contrast    
  #Creating the list of dataframe with all the contrast
  pair_res_list[[o_vs_o_contrast]] <- tt
}

#STEP 4 – Count, for each organ, in how many pairwise comparisons a gene is UP
#Creating an empty matrix "marker_counts"
all_genes <- rownames(log2cpm_matrix_limma)
marker_counts <- matrix(0, nrow = length(all_genes), ncol = length(organs), dimnames = list(all_genes, organs))

#With names() it iterates on the names of the list not the dataframes with that name
for (o_vs_o_contrast in names(pair_res_list)) {
  tt <- pair_res_list[[o_vs_o_contrast]] #extract that dataframe, then into tt
  
  # parse "SmallIntestine_vs_Liver"
  sp <- strsplit(o_vs_o_contrast, "_vs_")[[1]]
  organ_left  <- sp[1]   # e.g. "SmallIntestine"
  organ_right <- sp[2]   # e.g. "Liver"
  
  # gene significantly UP in 'organ_left' vs 'organ_right'
  sig_up <- tt$adj.P.Val < 0.1 & tt$logFC > 0.5
  up_genes <- tt$gene[sig_up]
  
  # add +1 to that organ for all genes that are up
  marker_counts[up_genes, organ_left] <- marker_counts[up_genes, organ_left] + 1
}

#STEP 5 – Select organ-specific marker genes (up in ≥5 other organs: threshold original paper with 11 organs)
marker_genes_by_organ <- list()

for (org in organs) {

  # Special condition for intestine 
  if (org %in% c("SmallIntestine", "LargeIntestine")) {
    min_hits <- 3
  } else {
    min_hits <- 4
  }

  idx <- marker_counts[, org] >= min_hits
  marker_genes_by_organ[[org]] <- rownames(marker_counts)[idx]
}

#lenght of significant (padj) marker gene found per organ
sapply(marker_genes_by_organ, length)

```

Plotting all marker gene per organ in a heatmap This code create an
heatmap with this pipeline: 1) Take the number of the shortest list of
marker genes among organs from marker_genes_by_organ --\> block lenght
2) From each other organ , from the list of marker genes, take the first
"block lenght" most upregulated genes (based on z-score of Log2(CPM) )
3) Outp: Plot a heatmap that shows for each organs the most
differentialy upregulated genes (marker/representative): 8 blocks of
"block lenght"

```{r}

suppressPackageStartupMessages({
  library(circlize)
})

top_n <- min(sapply(marker_genes_by_organ, length))

top_markers_by_organ_z <- list()

for (org in organs) {
  cand <- marker_genes_by_organ[[org]]
  cand <- intersect(cand, rownames(z_scores_organ)) #safety check
  
  if (length(cand) == 0) next #check
  
  #Extracting only can genes from z_scores_organ
  z_org <- z_scores_organ[cand, org]
  
  #sorting decrescenting z score
  top_genes <- names(sort(z_org, decreasing = TRUE))[1:min(top_n, length(z_org))]
  
  #sorting alphabetically
  top_genes <- sort(top_genes)
  
  #saving the list for this organ
   top_markers_by_organ_z[[org]] <- top_genes
}

#Final vector
final_genes <- unlist(top_markers_by_organ_z, use.names = FALSE)

row_block <- rep(names(top_markers_by_organ_z), sapply(top_markers_by_organ_z, length))

library(pheatmap)

# Prepara l'annotation per le righe
row_annotation <- data.frame(Organ = row_block)
rownames(row_annotation) <- final_genes

pheatmap(
  z_scores_organ[final_genes, organs],
  color = colorRampPalette(c("blue", "white", "red"))(100),
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  show_rownames = TRUE,
  annotation_row = row_annotation,
  fontsize_row = 7,
  fontsize_col = 10
)
```

Similar heatmap as above but with squares, readable genes, better style,
saved in pdf. But a bit long as image :(

```{r}
#Image in pdf with squares
library(pheatmap)

row_annotation <- data.frame(Organ = row_block)
rownames(row_annotation) <- final_genes

n_genes  <- length(final_genes)
n_organs <- length(organs)

pdf("heatmap_marker_organs.pdf",
    width  = 2 + n_organs * 0.5,  
    height = 4 + n_genes  * 0.25) 


pheatmap(
  z_scores_organ[final_genes, organs],
  color          = colorRampPalette(c("blue", "white", "red"))(100),
  cluster_rows   = FALSE,
  cluster_cols   = FALSE,
  annotation_row = row_annotation,
  show_rownames  = TRUE,
  cellwidth      = 15,   #squares
  cellheight     = 15,
  fontsize_row   = 8,
  fontsize_col   = 10
)

dev.off()
```

GO of the marker genes of each organ 1) Retrieve GO Biological Process
terms for all the genes in the block of marker genes for each organ. 2)
Count how many genes in that block (organ) share each GO term 3) Keep
only GO terms shared by at least min_genes and add GO term descriptions.
4) Apply this process to each organ’s marker genes. Count how many genes
share each GO term.

```{r}
BiocManager::install("org.Mm.eg.db")
```

```{r}
BiocManager::install("GO.db")
```

```{r}
library(org.Mm.eg.db)
library(AnnotationDbi)
library(dplyr)
library(GO.db)


get_GO_common <- function(genes_symbol, min_genes = 2) {
  # SYMBOL → ENTREZ
  eg <- AnnotationDbi::select(org.Mm.eg.db,
                              keys    = genes_symbol,
                              keytype = "SYMBOL",
                              columns = "ENTREZID") |>
    distinct() |>
    filter(!is.na(ENTREZID))
  
  if (nrow(eg) == 0) return(NULL)
  
  # GO for each gene (only Biological Process)
  go_tab <- AnnotationDbi::select(org.Mm.eg.db,
                                  keys    = eg$ENTREZID,
                                  keytype = "ENTREZID",
                                  columns = c("SYMBOL","GO","ONTOLOGY")) |>
    filter(!is.na(GO), ONTOLOGY == "BP")

  if (nrow(go_tab) == 0) return(NULL)
  
  # Collapse per GO: how may times it appears, and in which genes
  go_common <- go_tab |>
    group_by(GO) |>
    summarise(
      n_genes = n_distinct(SYMBOL),
      genes   = paste(sort(unique(SYMBOL)), collapse = ", "),
      .groups = "drop"
    ) |>
    arrange(desc(n_genes)) |>
    filter(n_genes >= min_genes)
  
  #Add description to GO
  go_common$term <- AnnotationDbi::Term(go_common$GO)
  
  go_common
  
}

#Apply to all organs
GO_common_by_organ <- lapply(top_markers_by_organ_z, get_GO_common)

#It returns the number of GO terms found for each organ.
sapply(GO_common_by_organ, nrow)

#See the GO with at least "min_genes" genes in a specific organ
GO_common_by_organ[["Heart"]]
```

Fig. 2b (1/2) Calculate "log2_rpkm" matrix from "counts_matrix_filtered"

```{r}

suppressPackageStartupMessages({
  library(edgeR)
  library(biomaRt)
})

# Gets all gene symbols as a character vector
mgi_ids <- rownames(counts_matrix_filtered)

# Establishes connection to mouse genome database
ensembl <- useMart("ensembl", dataset = "mmusculus_gene_ensembl")

# Calculate gene length using lenght of many transcript for each gene
len_tbl <- getBM(
  attributes = c("mgi_symbol", "ensembl_transcript_id", "transcript_length"),
  filters    = "mgi_symbol",
  values     = mgi_ids,
  mart       = ensembl
)

# Take the longest transcript length per gene
gene_len_tbl <- len_tbl %>%
  group_by(mgi_symbol) %>%
  summarise(gene_length_bp = max(transcript_length, na.rm = TRUE), .groups = "drop")

# Match lengths to original gene order
gene_len_bp <- gene_len_tbl$gene_length_bp[match(mgi_ids, gene_len_tbl$mgi_symbol)]
names(gene_len_bp) <- mgi_ids  #Name the vector with gene symbols

#Keep only genes with valid, positive lengths
keep <- !is.na(gene_len_bp) 

counts_for_rpkm <- counts_matrix_filtered[keep, ] # Keep only genes with the known lenght in database
gene_len_kb <- gene_len_bp[keep] / 1000 # Convert base pairs to kilobases


y <- DGEList(counts = counts_for_rpkm)
rpkm_matrix <- rpkm(y, gene.length = gene_len_kb)

log2_rpkm <- log2(rpkm_matrix + 0.01)

#Save the file
save(log2_rpkm, file = "log2_rpkm.RData")
load("log2_rpkm.RData")

```

Fig. 2b (2/2) 1) Taking the list of immunitary genes used by the authors
2) Use the raw log2(RPKM) as authors did to plot an heatmap of different
expression levels of this genes between the organs 3) Saving two
heatmap: 1. Auto-Scaled Heatmap (from min. to max. value from data) and
2. Fixed-Scale Heatmap (−5 to 15) p.s.: here between replicates it has
been done the mean of log2(RPKM)

```{r}
suppressPackageStartupMessages({library(pheatmap)
library(RColorBrewer)})

#Immuno genes from fig. 2b of paper
immunogenes_fig2b <- c("Ager","Slc40a1","Csf1r","Il10ra","Igsf21","Il27ra","Bmp6","Bmp7","Cpn","Fgfr2","Cd177","Sema6a","Ttr","Il1r",
                       "S100a8","S100a9","Cp","Adrb2","Cxc110","Dkk2","Il1r2","Vpr1","Map2","Argr2","Muc2","Lrp2","Lpl","Vegfa","Mdk",
                       "Sema4d","Ptn","Sema3c","Scarf1","Ltb","Clcf1","Nlrc5","Cc125","Sgpl1","Itga2b","Tnc","Itga3","Jag2","Sema7a",
                       "Erbb2","Lamc2","Met","Itgblb","Plau","P2ry12","Ccr5","Col12a1","Col5a1","Tbx2r","Ptnu","Col4a3","Col4a4",
                       "Col6a5","C3","Col5a1","Aib","Hic","Salipg","Sdcf","Alcam")

#org contains the vector of organs in the same order as log2_rpkm columns
org <- col_data$Organ[match(colnames(log2_rpkm), col_data$Run)]

#For each organ, calculates the mean expression of all genes across replicates
#Output: matrix where each column is an organ and each row is the mean log2-RPKM of a gene
mean_by_organ <- sapply(split(seq_along(org), org),
                        function(i) rowMeans(log2_rpkm[, i, drop = FALSE]))

#convert the matrix into a R dataframe
mean_by_organ_log2RPKM <- as.data.frame(mean_by_organ)

#In order to use z-scores of log2(RPKM)
#z_scores_organ <- t(scale(t(mean_by_organ))) to use z-score of log2(RKPM)

# Get the subset of genes that exist in both your 'genes' list and the mean_by_organ dataframe
genes_use <- intersect(immunogenes_fig2b, rownames(mean_by_organ_log2RPKM))

# Extract only those common genes from the mean_by_organ dataframe
matrix_RPKM_immunogenes <- mean_by_organ_log2RPKM[genes_use, , drop = FALSE]

keep <- rowSums(is.finite(as.matrix(matrix_RPKM_immunogenes))) > 0 #safety check
matrix_RPKM_immunogenes <- matrix_RPKM_immunogenes[keep, , drop = FALSE]

#In order to choose the right order of the organs and rename it
matrix_RPKM_immunogenes <- matrix_RPKM_immunogenes[, c("Brain","Heart","Kidney","LargeIntestine","Liver","Lung","SmallIntestine", "Spleen"), drop = FALSE]

colnames(matrix_RPKM_immunogenes) <- c("Brain","Heart","Kidney","Large int.","Liver","Lung","Large int.", "Spleen")

#heatmap

# fixed breaks from -5 to 15
bk <- seq(-5, 15, length.out = 51)

# same white/yellow → orange → dark red gradient
cols <- colorRampPalette(brewer.pal(9, "YlOrRd"))(50)


#auto scale (the more correct one but turns out diff. max. value from paper)
pheatmap(
  matrix_RPKM_immunogenes,
  color = colorRampPalette(brewer.pal(9, "YlOrRd"))(50),
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  filename     = "fig2b_autoscale.pdf",
  border_color = "grey90",
  cellwidth  = 12,
  cellheight = 12
)

#fixed scale (all value above 15 will be the same color)
pheatmap(
  matrix_RPKM_immunogenes,
  color        = cols,
  breaks       = bk,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_row = 7,
  fontsize_col = 9,
  border_color = "grey90",          
  cellwidth  = 12,  
  cellheight = 12,   
  filename     = "fig2b_fixedscale.pdf",
  legend_breaks  = c(-5, 0, 5, 10, 15),
  legend_labels  = c("-5", "0", "5", "10", "15")
)
```

Fig2b - Alternative averaging p.s.: here between replicates it has been
done the log2(mean RPKM)

```{r}
#try to do log2(mean RPKM) just to check --> it also worst

#mean RPKM per organ (no log yet)
mean_rpkm_by_organ <- sapply(split(seq_along(org), org),
                             function(i) rowMeans(rpkm_matrix[, i, drop = FALSE]))

#now take log2(mean RPKM + 1)
log2_mean_rpkm_by_organ <- log2(mean_rpkm_by_organ + 1)

log2_mean_rpkm_by_organ <- as.data.frame(log2_mean_rpkm_by_organ)

genes_use2 <- immunogenes_fig2b[immunogenes_fig2b %in% rownames(log2_mean_rpkm_by_organ)]
matrix_logmean_immunogenes <- log2_mean_rpkm_by_organ[genes_use2, , drop = FALSE]

keep2 <- rowSums(is.finite(as.matrix(matrix_logmean_immunogenes))) > 0
matrix_logmean_immunogenes <- matrix_logmean_immunogenes[keep2, , drop = FALSE]

# Same organ order and same names as before
matrix_logmean_immunogenes <- matrix_logmean_immunogenes[, c("Brain","Heart","Kidney","LargeIntestine","Liver","Lung","SmallIntestine", "Spleen"), drop = FALSE]

colnames(matrix_logmean_immunogenes) <- c("Brain","Heart","Kidney","Large int.","Liver","Lung","Large int.", "Spleen")

# fixed breaks from -5 to 15
bk <- seq(-5, 15, length.out = 101)
cols <- colorRampPalette(brewer.pal(9, "YlOrRd"))(100)

#auto scale (the more correct one but turns out diff. max. value from paper)
pheatmap(
  matrix_logmean_immunogenes,
  color = colorRampPalette(brewer.pal(9, "YlOrRd"))(50),
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  filename     = "fig2b_autoscale_logmean.pdf",
  border_color = "grey90",
  cellwidth  = 12,
  cellheight = 12
)

#fixed scale (all value above 15 will be the same color)
pheatmap(
  matrix_logmean_immunogenes,
  color        = cols,
  breaks       = bk,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_row = 7,
  fontsize_col = 9,
  border_color = "grey90",          
  cellwidth  = 12,  
  cellheight = 12,   
  filename     = "fig2b_fixedscale_logmean.pdf",
  legend_breaks  = c(-5, 0, 5, 10, 15),
  legend_labels  = c("-5", "0", "5", "10", "15")
)
```

Fig. 2c 1.Start from log-transformed gene expression (log₂RPKM).
2.Compute the mean expression of each gene within each organ (btwn
replicates). mean(log₂RPKM) 3.Normalize each gene across organs using
gene-wise z-scores. z score of mean(log₂RPKM) 4.Match each gene to its
assigned module (TNF, IFN, chemokines, etc.) from suppl4
(supplementary_table_4 of paper) 5.Compute each module’s signature as
the average z-score of its genes per organ. 6.Plot modules across organs
using signature intensity (mean of genes z-scores of that module in that
organ)

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)})

#cleaning the csv from the paper 44.csv (supplemetanry_table_4.csv)

# 1. READ the original file with the correct separator
suppl4_clean <- read.csv("supplementary_table_4.csv", sep = ",", header = TRUE, stringsAsFactors = FALSE)
colnames(suppl4_clean) <- c("Gene", "Group", "Type")  # Rinominato come "Gene"
```

```{r}
# 3. REMOVE empty rows – FIX: use "Gene" not "Gene Symbol"
suppl4_clean <- suppl4_clean[!is.na(suppl4_clean$Gene) & suppl4_clean$Gene != "", ]

# 4. CHECK
print("Data from supplementary table:")
print(head(suppl4_clean))
print(paste("Number of genes:", nrow(suppl4_clean)))

#5. SAVE as standard CSV
write.csv(suppl4_clean, "suppl4_right.csv", row.names = FALSE)
print("File saved as 'suppl4_right.csv'")

suppl4 <- suppl4_clean

#Code
save(suppl4, file = "suppl4_right.RData")

load("suppl4_right.RData")

## Mean expression per gene × organ without pivoting to long format

# 1) numeric matrix: genes x runs
mat <- as.matrix(log2_rpkm)   
storage.mode(mat) <- "numeric"


# 2) organ label for each run (aligned to columns of mat)
runs  <- colnames(mat)
organ <- col_data$Organ[ match(runs, col_data$Run) ]

# 3) sum expression per organ (rowsum works on rows → transpose first)
sum_by_organ <- rowsum(t(mat), group = organ)  # organs x genes

# 4) convert sums to means
n_per_organ   <- table(organ)[rownames(sum_by_organ)]
mean_by_organ <- t( sweep(sum_by_organ, 1, n_per_organ, "/") )  # genes x organs

# 5) final data.frame: gene × organ
expr_mat <- data.frame(
  gene = rownames(mean_by_organ),
  mean_by_organ,
  check.names = FALSE
)

rownames(expr_mat) <- expr_mat$gene
expr_mat$gene <- NULL

expr_norm <- t(scale(t(expr_mat)))   # gene-wise z-score

modules <- suppl4 %>%
  dplyr::rename(gene = `Gene`, 
                module = Group, 
                type = Type)

modules$module_full <- paste0(modules$module, " (", substr(modules$type,1,1), ")")

# Filter genes that are in your expression matrix
modules_use <- modules %>% filter(gene %in% rownames(expr_norm))

# For each module, average normalized expression across genes
module_signatures <- modules_use %>%
  group_by(module_full) %>%
  do({
    genes <- .$gene
    data.frame(t(colMeans(expr_norm[genes, , drop=FALSE])))
  }) %>%
  ungroup()

sig_mat <- as.data.frame(module_signatures)
rownames(sig_mat) <- sig_mat$module_full
sig_mat$module_full <- NULL

library(reshape2)
library(ggplot2)

df_plot <- melt(as.matrix(sig_mat))
colnames(df_plot) <- c("module", "organ", "value")

ggplot(df_plot, aes(x=organ, y=module, color=value, size=value)) +
  geom_point() +
  scale_color_gradient2(low="yellow", mid="orange", high="red") +
  theme_bw()

```

-- NOT NECESSARY-- Fig. 2c The old method uses z-scores of
mean(log2(RPKM)) while this only use mean(Log2(RPKM)) of the genes in a
module and keep them (plotting mean log2RPKM of the module) Just
enphasize the color but doesn't change the relative expressions, I just
wanted to see if it changes

```{r}
#Z SCORE after log2RPKM
load("suppl4_right.RData")

## 1) Mean expression per gene × organ without pivoting to long format

# Convert to numeric matrix WITHOUT reintroducing "gene" column
mat <- as.matrix(log2_rpkm)   
storage.mode(mat) <- "numeric"

runs  <- colnames(mat)
organ <- col_data$Organ[ match(runs, col_data$Run) ]

# Remove samples with NA organ
keep <- !is.na(organ)
mat   <- mat[, keep, drop = FALSE]
runs  <- runs[keep]
organ <- organ[keep]

# Now aggregate
sum_by_organ  <- rowsum(t(mat), group = organ)  # organs x genes
n_per_organ   <- table(organ)[rownames(sum_by_organ)]
mean_by_organ <- t( sweep(sum_by_organ, 1, n_per_organ, "/") )  # genes x organs

expr_mat <- data.frame(
  gene = rownames(mean_by_organ),
  mean_by_organ,
  check.names = FALSE
)
rownames(expr_mat) <- expr_mat$gene
expr_mat$gene <- NULL   # now: genes x organs, mean log2RPKM

## 2) Modules
modules <- suppl4 %>%
  dplyr::rename(gene = `Gene`, 
                module = Group, 
                type = Type)

modules$module_full <- paste0(modules$module, " (", substr(modules$type,1,1), ")")

modules_use <- modules %>% filter(gene %in% rownames(expr_mat))

## 3) For each module, average *log2RPKM* across genes
module_signatures <- modules_use %>%
  group_by(module_full) %>%
  do({
    genes <- .$gene
    data.frame(t(colMeans(expr_mat[genes, , drop=FALSE])))
  }) %>%
  ungroup()

sig_mat <- as.data.frame(module_signatures)
rownames(sig_mat) <- sig_mat$module_full
sig_mat$module_full <- NULL    # modules x organs, mean log2RPKM

## 4A) OPTION 1: z-score per module (row-wise)
sig_z <- t(scale(t(sig_mat)))  # module-wise z-score

## 4B) OPTION 2: 0–1 scaling per module (often closer to “Low–High” legend)
# sig_z <- t(apply(sig_mat, 1, function(x) (x - min(x)) / (max(x) - min(x))))

library(reshape2)
library(ggplot2)

df_plot <- melt(as.matrix(sig_z))
colnames(df_plot) <- c("module", "organ", "value")

ggplot(df_plot, aes(x=organ, y=module, color=value, size=value)) +
  geom_point() +
  scale_color_gradient(low="lightyellow", high="red") +
  theme_bw()

```

Plot of Fig. 2c with a better professional and organized style
(auto-scaled)

```{r}
library(ggplot2)
library(reshape2)

## ---- Prepare data ----
df_plot <- melt(as.matrix(sig_z))
colnames(df_plot) <- c("module", "organ", "value")

# Order modules (optional)
df_plot$module <- factor(
  df_plot$module,
  levels = rev(sort(unique(df_plot$module)))
)

## ---- Define breaks you want in the legend ----
my_breaks <- seq(-2, 2, 1)   # -2 -1 0 1 2

my_cols <- colorRampPalette(brewer.pal(9, "YlOrRd"))(50)

## ---- Build plot ----
p <- ggplot(df_plot, aes(x = organ, y = module)) +
  geom_point(
    aes(colour = value, size = value),
    shape = 16,
    alpha = 0.95
  ) +
  
    #color scale
    scale_colour_gradientn(
    colours = my_cols,
    limits = c(min(df_plot$value), 2),
    breaks = my_breaks,
    name = "Signature intensity"
  ) +
  
  # SIZE SCALE (same breaks!)
  scale_size(
    range  = c(0.2, 2),
    breaks = my_breaks,
    guide  = "none"      # hide size legend, we will override through color legend
  ) +
  
  # LEGEND: override the size of the points
  guides(
    colour = guide_legend(
      override.aes = list(
        size = scales::rescale(my_breaks, to = c(0.2, 2))
      ),
      title.position = "top",
      direction      = "horizontal"
    )
  ) +
  
  ylab(NULL) +
  xlab(NULL) +
  coord_equal() +
  theme_bw(base_size = 9) +
  theme(
    axis.text.x      = element_text(angle = 90, vjust = 0.5, hjust = 1, colour="black"),
    axis.text.y      = element_text(colour="black"),
    axis.title.x     = element_text(colour="black"),
    axis.title.y     = element_text(colour="black"),
    legend.text      = element_text(colour="black"),
    legend.title     = element_text(colour="black"),
    panel.grid.major = element_line(size = 0.3, colour = "grey90"),
    panel.grid.minor = element_blank(),
    panel.border     = element_rect(colour = "black", fill = NA),
    legend.position  = "bottom",
    legend.box.margin = margin(t = 5, r = 5, b = 20, l = 5)
  )

p

## ---- SAVE HIGH-QUALITY PDF ----
ggsave("Fig2c.pdf", p,
       width = 10, height = 8, units = "in",
       dpi = 300)
```

## 8. Grouping of identified DEGs

A key decision is how to select the differentially expressed genes
(based on ANOVA, DESeq2, combination of the two, just the immune systems
related ones), perhaps the 144 seen above, which would be a combination
of ANOVA and DESeq2 (against liver).

```{r}
# Extract the list of DE genes
DEG <- res_DE$gene[res_DE$DE]
```
     
```{r}
# Calculate the average expression for each gene across different organs
avg_expr <-
  sapply(
    sort(unique(meta$Organ)),
    function(Organ) {
      rowMeans(counts_matrix_filtered[, meta$Organ == Organ])
    }
  )

organs <- sort(unique(meta$Organ))
colnames(avg_expr) <- organs
```


```{r}
# Identify the layer in which each DE gene has the highest average expression
max_organ_DEG <- setNames(
  colnames(avg_expr)[apply(avg_expr[DEG, , drop = FALSE], 1, which.max)],
  DEG
)


## Frequency table of max organ per DEG
table(max_organ_DEG)

## Split DEGs by their max organ and extract avg_expr rows
avg_expr_DEG_list <- tapply(
  names(max_organ_DEG),
  max_organ_DEG,
  function(x) avg_expr[x, , drop = FALSE]
)

## Scale each submatrix by gene (row)
scaled_expr_DEG_list <- lapply(
  avg_expr_DEG_list,
  function(x) t(scale(t(x)))
)

## Layout for 8 boxplots (adjust if needed)
layout(matrix(1:8, nrow = 2, byrow = TRUE))
par(mar = c(3, 3, 3, 3))

## One boxplot per organ
for (organ in names(scaled_expr_DEG_list)) {
  mat <- scaled_expr_DEG_list[[organ]]
  boxplot(
    mat,
    main = paste0(organ, " (", nrow(mat), ")")
  )
}

```
```{r}
typeof(expr)   # should be "matrix" or "double"
class(expr) 
```
```{r}
# Compute average expression per organ
avg_expr <- sapply(
  sort(unique(meta$Organ)),
  function(Organ) {
    rowMeans(counts_matrix_filtered[, which(meta$Organ == Organ)])
  }
)
```

```{r}
# Ensure it's a matrix and assign column names
avg_expr <- as.matrix(avg_expr)
colnames(avg_expr) <- organs

# Extract DE genes
DE_genes <- res_DE$gene[res_DE$DE]

# Compute Spearman correlation among average expression of DE genes across organs
corr_DEG <- cor(
  avg_expr[DE_genes, , drop = FALSE],
  method = "spearman"
)

# Hierarchical clustering
hcl_DEG <- hclust(
  as.dist(1 - corr_DEG),
  method = "complete"
)

# Plot dendrogram without labels
plot(hcl_DEG, labels = FALSE)

```


```{r}
library(gplots)

heatmap.2(
  corr_DEG,
  Rowv = as.dendrogram(hcl_DEG),
  Colv = as.dendrogram(hcl_DEG),
  trace = "none",
  scale = "none",
  labRow = organs,
  labCol = organs,
)
```

2025.11.23 working on this section

### 99. Further work

In addition the three differential expression calculation methods used:
ANOVA, DESeq2 and Limma, ANCOVA based on p-value and padj was also
attempted to generate vulcano plots. However, the plots are not
satisfactory.

VOLCANO ANCOVA with padj - one organ vs all

```{r}

#M: This should be ANCOVA i'm trying to have a vulcano. I asked gpt because I haven't found the code on the tutorial.
#padj < 0.05 is severe so it is possible that there aren't gene in the replicates that satisfyes all the condition
#I'm trying eith 0.1 of padj
meta <- col_data[ match(colnames(counts_matrix_filtered), col_data$Run), ]

organ_of_interest <- "Liver"

# 1. Subset samples
group_A <- meta$Organ == organ_of_interest
group_B <- meta$Organ != organ_of_interest

expr_A <- counts_matrix_filtered[, group_A]
expr_B <- counts_matrix_filtered[, group_B]

# 2. Compute log2 fold-change
log2FC <- log2(rowMeans(expr_A + 1) / rowMeans(expr_B + 1))

# 3. Compute p-values (t-test)
pvals <- apply(counts_matrix_filtered, 1, function(e){
  t.test(e[group_A], e[group_B])$p.value
})

# 4. Adjust p-values
padj <- p.adjust(pvals, method = "BH")

# 5. Create results table
suppressPackageStartupMessages({library(dplyr) 
library(ggplot2)
library(ggrepel)})

res_DE <- res_DE %>% 
  mutate(
    log2FC = log2FC,
    padj   = padj,
    status = case_when(
  padj < 0.1 & log2FC >  1 ~ "Up",
  padj < 0.1 & log2FC < -1 ~ "Down",
  TRUE                      ~ "NotSig"
),
    label = if_else(padj < 0.1, gene, NA_character_)  # label only sig if you want
  )


#plot volcano
ggplot(res_DE, aes(x = log2FC, y = -log10(padj),
                color = status, label = label)) +
  geom_point() +
  geom_text_repel(max.overlaps = 15) +
  geom_vline(xintercept = c(-1, 1), linetype = "dotted") +
  geom_hline(yintercept = -log10(0.1), linetype = "dotted") +
  scale_color_manual(values = c(
    "Down"   = "blue",
    "NotSig" = "grey80",
    "Up"     = "red"
  )) +
  theme_minimal()
```

VOLCANO ANCOVA with pvalue - one organ vs all

```{r}
#Same ANCONA volcono but with pvalue (0.05) and not padj(0.1)

organ_of_interest <- "Liver"

# 1. Subset samples
group_A <- meta$Organ == organ_of_interest
group_B <- meta$Organ != organ_of_interest

expr_A <- counts_matrix_filtered[, group_A]
expr_B <- counts_matrix_filtered[, group_B]

# 2. Compute log2 fold-change
log2FC <- log2(rowMeans(expr_A + 1) / rowMeans(expr_B + 1))

# 3. Compute p-values (t-test)
pvals <- apply(counts_matrix_filtered, 1, function(e){
  t.test(e[group_A], e[group_B])$p.value
})

suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(ggrepel)
})

# 4. Create results table (usando pval, non padj)
res_DE <- res_DE %>% 
  mutate(
    log2FC = log2FC,
    pval   = pvals,
    status = case_when(
      pval < 0.05 & log2FC >  1 ~ "Up",
      pval < 0.05 & log2FC < -1 ~ "Down",
      TRUE                      ~ "NotSig"
    ),
    label = if_else(pval < 0.05, gene, NA_character_)  # etichette sui p<0.05
  )

# volcano con p-value
ggplot(res_DE, aes(x = log2FC, y = -log10(pval),
                color = status, label = label)) +
  geom_point() +
  geom_text_repel(max.overlaps = 15, na.rm = TRUE) +
  geom_vline(xintercept = c(-1, 1), linetype = "dotted") +
  geom_hline(yintercept = -log10(0.05), linetype = "dotted") +
  scale_color_manual(values = c(
    "Down"   = "blue",
    "NotSig" = "grey80",
    "Up"     = "red"
  )) +
  theme_minimal()
```