---
title: "4_differential_gex_ANOVA_and_DESeq2"
author: "Marco Fiorani, Sara Leka and Carlo Rossetto"
date: "2025-11-30"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(dplyr)
  library(biomaRt)
  library(sva)
  library(pheatmap) 
  library(ggplot2)
  library(gridExtra)
})
```

For differential expression, three approaches were used: ANOVA (as per paper), DESeq2 (as per tutorial) and Limma (as per paper). In this script the first two approaches are shown.

### Approach 1: ANOVA, our own method as per tutorial

```{r}
col_data = meta_data[c("Run","Organ")]
DE_test <- function(expr,
                    cond,
                    ctrl = NULL,
                    covar = NULL,
                    padj_method = p.adjust.methods) {
  
  # Apply the test gene-by-gene
  pval_fc <- data.frame(
    t(
      pbapply(expr, 1, function(e) {
        
        # Build data frame for the model
        dat <- data.frame(
          y = log1p(e),   # log-transformed raw counts
          cond = cond     # condition/group variable
        )
        
        # Add covariates if provided
        if (!is.null(covar)) {
          dat <- data.frame(dat, covar)
        }
        
        # Full model: y ~ cond + covariates
        m1 <- lm(y ~ ., data = dat)
        
        # Reduced model: y ~ covariates (without cond)
        m0 <- lm(y ~ . - cond, data = dat)
        
        # ANOVA test comparing full vs reduced model
        test <- anova(m1, m0)
        pval <- test$Pr[2]   # p-value for the effect of cond
        
        # Mean log-expression per condition
        avgs <- tapply(log1p(e), cond, mean)
        
        # Fold-change calculation
        if (!is.null(ctrl) && sum(cond %in% ctrl) > 0) {
          # FC vs reference control group
          fc <- exp(max(avgs[names(avgs) != ctrl]) - avgs[ctrl])
        } else {
          # FC = max group vs min group
          fc <- exp(max(avgs) - min(avgs))
        }
        
        return(c(pval = unname(pval), fc = unname(fc)))
      })
    ),
    row.names = rownames(expr)
  )
  
  # Multiple-testing correction
  padj <- p.adjust(pval_fc$pval, method = padj_method)
  
  # Return result table
  return(data.frame(pval_fc, padj = padj)[, c("pval", "padj", "fc")])
}
```

```{r}
suppressPackageStartupMessages({library(pbapply)}) 

res_DE <- DE_test(
  expr  = counts_matrix_filtered,
  cond  = meta_data$Organ,
  covar = NULL
) %>%
  tibble::rownames_to_column("gene")
```

```{r}
suppressPackageStartupMessages({library(pbapply)}) 

res_DE_contamination <- DE_test(
  expr  = counts_matrix_filtered,
  cond  = meta_data$Organ,
  covar = meta_data$ContaminationScore
) %>%
  tibble::rownames_to_column("gene")
```


```{r}
# 2. Create DE flags (MANDATORY)
res_DE <- res_DE %>%
  mutate(DE  = padj < 0.05 & fc > 2) %>%   # TRUE/FALSE
  mutate(DEG = ifelse(DE, gene, NA))      # label only DE genes
```

```{r}
suppressPackageStartupMessages({library(ggrepel)})
ggplot(res_DE, aes(x = log(fc), y = -log10(padj), col=DE, label=DEG)) +
  geom_point() +
  geom_text_repel() +
  geom_vline(xintercept=c(log(2), 0), col="#303030", linetype="dotted") +
  geom_hline(yintercept=-log10(0.1), col="#303030", linetype="dotted") +
  scale_color_manual(values=c("#909090", "red")) +
  theme_minimal()
```

```{r}
library(ggplot2)
library(dplyr)
library(ggpubr)

# Combine results
res_combined <- bind_rows(
  res_DE %>% mutate(method = "No Covariate"),
  res_DE_contamination %>% mutate(method = "With Contam. Covar")
)

# 1. Volcano plot overlay
p1 <- ggplot(res_combined, aes(x = log(fc), y = -log10(padj), color = method)) +
  geom_point(alpha = 0.6, size = 1) +
  facet_wrap(~method) +
  labs(title = "Volcano Plot Comparison") +
  theme_bw()

print(p1)
```

```{r}
# 2. P-value difference plot
res_diff <- inner_join(
  res_DE %>% dplyr::select(gene, pval_no_covar = pval, fc_no_covar = fc, DE),
  res_DE_contamination %>% dplyr::select(gene, pval_contam = pval, fc_contam = fc),
  by = "gene"
) %>%
  mutate(pval_diff = pval_no_covar - pval_contam)

p2 <- ggplot(res_diff, aes(x = fc_no_covar, y = pval_diff, color = DE)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_x_log10() +
  scale_color_manual(values = c("FALSE" = "grey50", "TRUE" = "red"), 
                     name = "DE (padj<0.05,\nFC>2)", 
                     labels = c("No", "Yes")) +
  labs(x = "FC (No Covar, log10)", y = "Δp-value (No Covar - Contam)", 
  title = "P-value shift by contamination correction") +
  theme_bw()

# same plot only with only the differentially expressed genes and the ones that lose their significance after contamination

# Filter to ONLY DE genes for the second plot
res_diff_DE <- res_diff %>%
  filter(DE == TRUE)  # Keep only DE genes (padj < 0.05 & fc > 2)
```

```{r}
# EXPLICIT: Only genes that were DE in res_DE (no covariate)
res_diff_DE_only <- inner_join(
  res_DE %>% 
    filter(DE == TRUE) %>%  # ONLY DE genes from res_DE
    dplyr::select(gene, pval_no_covar = pval, fc_no_covar = fc, padj_no_covar = padj),
  res_DE_contamination %>% 
    dplyr::select(gene, pval_contam = pval, fc_contam = fc, padj_contam = padj),
  by = "gene"
) %>%
  mutate(
    pval_diff = pval_no_covar - pval_contam,
    lost_sig = padj_no_covar < 0.05 & padj_contam >= 0.05
  )

# Plot: ONLY DE genes from res_DE
p2_DE_only <- ggplot(res_diff_DE_only, aes(x = fc_no_covar, y = pval_diff, color = lost_sig)) +
  geom_point(alpha = 0.6, size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  scale_x_log10() +
  scale_color_manual(
    values = c("FALSE" = "red", "TRUE" = "orange"), 
    name = "Lost sig\nafter contam",
    labels = c("Remains DE", "Lost DE")
  ) +
  labs(x = "FC (No Covar, log10)", y = "Δp-value", 
       title = "ONLY genes DE in res_DE (no covariate)") +
  theme_bw()

print(p2)
print(p2_DE_only)
```

```{r}
# 3. Bland-Altman for padj (SIMPLE approach)
res_padj <- inner_join(
  res_DE %>% dplyr::select(gene, padj_no_covar = padj),
  res_DE_contamination %>% dplyr::select(gene, padj_contam = padj),
  by = "gene"
) %>%
  mutate(
    mean_padj = (padj_no_covar + padj_contam)/2,
    diff_padj = padj_no_covar - padj_contam,
    lost_sig = padj_no_covar < 0.05 & padj_contam >= 0.05  # Sig before, NOT after
  )

p3 <- ggplot(res_padj, aes(x = padj_no_covar, y = diff_padj, color = lost_sig)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(alpha = 0.6, size = 1) +
  scale_x_log10() +
  scale_color_manual(
    values = c("FALSE" = "blue", "TRUE" = "orange"), 
    name = "Status after contam.",
    labels = c("Sig before (various)", "Sig→NOT sig")
  ) +
  labs(x = "Mean padj (log10)", y = "Δpadj (No Covar - Contam)", 
       title = "Bland-Altman: padj agreement") +
  theme_bw()

print(p3)
```

```{r}
# 4. Summary table
summary_stats <- res_combined %>%
  mutate(sig = padj < 0.05) %>%
  count(method, sig)

print("DE genes summary:")
print(summary_stats)

```
The contamination seems significant!

```{r}
# Get genes that were DE before (res_DE) but NOT after contamination correction
lost_DE_genes <- inner_join(
  res_DE %>% dplyr::select(gene, padj_no_covar = padj),
  res_DE_contamination %>% dplyr::select(gene, padj_contam = padj),
  by = "gene"
) %>%
  filter(padj_no_covar < 0.05 & padj_contam >= 0.05) %>%  # Sig before, NOT after
  pull(gene)

# Output the list
print("Genes that lost significance after contamination correction:")
print(lost_DE_genes)

# Optional: Save to file
write.csv(data.frame(gene = lost_DE_genes), 
          "lost_DE_genes_contamination.csv", 
          row.names = FALSE)

# Count how many
cat("Total genes that lost DE status:", length(lost_DE_genes), "\n")
```
Adding the covariance does not influence the fold change 

Note the approach above finds the maximum and minimum count amongst organs and computes one fold change per gene. So one gene fold change may be: 
max gene count from sample in liver - max gene count from sample in brain, while another may be: max gene count from sample in kidney - max gene count from sample in small intestine, there is one "fold change" value per gene.

Also the fold change is linear, not log(base 2) (?).

### Approach 2: DESeq2

#### Calculating the Log2fold changes

```{r}
suppressPackageStartupMessages(library(DESeq2))

# preparing the tables (using the counts_matrix_filtered as recommended)
# converting the counts to integers (by rounding them) and the table type to matrix
samples_mean = rowMeans(counts_matrix_filtered)
full_cmf = cbind(counts_matrix_filtered,samples_mean)
```

```{r}
# Build col_data PROPERLY as data.frame
col_data <- meta_data[c("Run","Organ","ContaminationScore")]
rownames(col_data) <- col_data$Run  # Set rownames FIRST

# Add dummy row as data.frame (NOT vector with rbind!)
dummy_row <- data.frame(
  Run = "samples_mean",
  Organ = "organs mean", 
  ContaminationScore = 0,  # Numeric!
  row.names = "samples_mean"
)

col_data <- rbind(col_data, dummy_row)  # Now both are data.frames
col_data$ContaminationScore <- as.numeric(col_data$ContaminationScore)  # Ensure numeric

# Factor Organ properly
col_data$Organ <- factor(col_data$Organ)
col_data$Organ <- relevel(col_data$Organ, ref = "organs mean")
```

```{r}
# generating the differential expression data
dds_filtered <-  DESeqDataSetFromMatrix(countData = rounded_cmf, colData = col_data,design = ~ Organ)
dds_filtered <- DESeq(dds_filtered, test="LRT", reduced = ~1)
res_DESeq2 <- results(dds_filtered)
```
creating a table with the top 10 differentially expressed genes per organ

```{r}
# Get all organs except the reference "organs mean"
organs <- setdiff(levels(col_data$Organ), "organs mean")

get_top10_for_organ <- function(org) {
  res <- results(dds_filtered,
                 contrast = c("Organ", org, "organs mean"))

  # convert to plain data.frame and add gene, organ columns
  df <- as.data.frame(res)
  df$gene  <- rownames(df)
  df$organ <- org

  # remove rows without log2FC
  df <- df[!is.na(df$log2FoldChange), ]

  # order by absolute log2FC (largest first)
  o <- order((df$log2FoldChange), decreasing = TRUE)
  df <- df[o, ]

  # keep top 10 rows (or fewer if <10 genes)
  if (nrow(df) > 10) df <- df[1:10, ]

  # keep only requested columns
  df[, c("gene", "log2FoldChange", "organ")]
}

# apply to each organ and combine
top10_list  <- lapply(organs, get_top10_for_organ)
top10_table <- do.call(rbind, top10_list)
```

```{r}
dds_filtered_con <-  DESeqDataSetFromMatrix(countData = rounded_cmf, colData = col_data,design = ~ ContaminationScore + Organ)
dds_filtered_con <- DESeq(dds_filtered_con, test="LRT", reduced = ~ContaminationScore)
res_DESeq2_con <- results(dds_filtered_con)
```

The default results(dds_filtered) after an LRT gives you this global
p‑value per gene, plus one particular log2FC column (often the first
non‑reference level vs reference), but the p‑value is for “any organ
difference,” not for that single pairwise contrast.

If you want specific organ vs organ log2FCs after fitting with LRT, you
still call results(dds_filtered, contrast = c("Organ", "Kidney",
"Liver")), which will use the already estimated coefficients to compute
that pairwise log2FC; the LRT p‑values remain the same global ones.

(1) key difference is that the fold changes, in comparison to the ANOVA
    above, are log base 2. (2) The log2fold changes are in comparison to
    one reference organ, and not max() - min() across all organs like in
    ANOVA. To use RE_Seq2 to calculate log2fold changes against a mean
    count of all organs one would either have to introduce an artificial
    factor/organ that is the calculated mean across all organs or
    calculate it manually by iterating through all organs and dividing
    by the averaged mean count for that gene.

Next is some visualisation of the DESeq 2 results, it is extra, not in
tutorial.

####Visualising 

This section is extra (not from tutorial), the section
"8. Grouping DEGs" has the same scope and follows the tutorial, so that one is better to follow.

This first plot is the log2fold change versus normalised gene count for two organs (average of samples across the organ) against the reference organ (its samples averaged). Really just for exploration.

```{r}
res <- results(dds_filtered, contrast = c("Organ", "SmallIntestine","LargeIntestine"))
plotMA(res, main="MA-plot:SmallIntestine and Liver vs Mean of Organs", ylim=c(-10,10))
legend("topright", legend=c("Small intestine", "Large intestine"), 
       col=c("blue", "grey"), pch=16, bty="n")
```

Visualising the normalised counts of the top 20 differentially expressed genes based on DESeq2.

```{r}
library(pheatmap)
vsd <- vst(dds_filtered, blind=FALSE)  # variance stabilizing transform
top_genes <- head(order(res$padj), 20)  # top 20 differentially expressed genes considering small intestine and large intestine versus reference organ, liver
norm_counts <- assay(vsd)[top_genes, ]
pheatmap(norm_counts, cluster_rows=TRUE, cluster_cols=TRUE,
         annotation_col=as.data.frame(colData(dds_filtered)["Organ"]),
         main="Normalised counts heatmap of top 20 most differentially expressed genes")
```

more DESeq2 plotting, now the log base2 fold changes of all the
differentially expressed genes (according to DESeq2)

```{r}
organ_levels <- levels(col_data$Organ)
organ_levels <- organ_levels[organ_levels != "organs mean"]  # exclude reference (the same organ that was set as reference three chunks ago)

# Create a matrix to hold log2 fold changes: genes x organs
log2fc_matrix <- sapply(organ_levels, function(org) {
  res_org <- results(dds_filtered, contrast=c("Organ", org, "organs mean"))
  log2fc <- res_org$log2FoldChange
  names(log2fc) <- rownames(res_org)
  return(log2fc)
})

# Convert to matrix and remove genes with NA values in any organ
log2fc_matrix <- as.matrix(log2fc_matrix)
log2fc_matrix <- log2fc_matrix[complete.cases(log2fc_matrix), ]

# Generate heatmap of log2 fold changes
pheatmap(log2fc_matrix,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames = FALSE,
         main = "Heatmap of log2 fold changes across organs compared to liver",
         fontsize_col = 12,
         color = colorRampPalette(c("blue", "white", "red"))(100))
```

same as last plot but only with top 20 most differentially expressed
genes acrossed all organs (the ones with the highest variance in
log2fold change amongs all organs against the reference organ):

```{r}
# Calculate variance for each gene (row)
gene_variances <- apply(log2fc_matrix, 1, var)

# Select top N genes by variance, for example top 20
top_genes <- order(gene_variances, decreasing = TRUE)[1:20]

# Subset the log2fc matrix to include only top variable genes
log2fc_top <- log2fc_matrix[top_genes, ]

# Now you can use 'log2fc_top' to plot the heatmap
library(pheatmap)
pheatmap(log2fc_top,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames = TRUE,
         main = "Log2Fold Change of Top 20 Most Differentially Expressed Genes Compared to Liver",
         legend=TRUE,
         annotation_legend=TRUE,
         color = colorRampPalette(c("blue", "white", "red"))(100))
```

Comparing DESeq2 with the ANOVA and ANCOVA gene expression: approach 1
vs approach 2

```{r}

cor(res_DESeq2$padj, res_DE %>% filter(gene %in% rownames(res_DESeq2)) %>% pull(padj), method = "spearman", use="complete.obs")

layout(matrix(1:2, nrow = 1))

## standard scatter plot
plot(
  -log10(res_DESeq2$pvalue),
  -log10(
    res_DE %>%
      dplyr::filter(gene %in% rownames(res_DESeq2)) %>%
      dplyr::pull(pval)
  ),
  xlab = "-log10(pval DESeq2)",
  ylab = "-log10(pval DE)",
  pch  = 16
)

## smooth density scatter plot
smoothScatter(
  -log10(res_DESeq2$pvalue),
  -log10(
    res_DE %>%
      dplyr::filter(gene %in% rownames(res_DESeq2)) %>%
      dplyr::pull(pval)
  ),
  xlab = "-log10(pval DESeq2)",
  ylab = "-log10(pval DE)",
  pch  = 16
)

```

```{r}
table(
  p.adjust(res_DESeq2$pvalue, method = "bonferroni") < 0.05,
  res_DE %>%
    dplyr::filter(gene %in% rownames(res_DESeq2)) %>%
    dplyr::pull(padj) < 0.05
)
```
